---
title: "02-ConservationAreas"
author: "Dimitrios Markou"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Module 2: Conservation Areas: KBA's and Priority Places

##### Author: Dimitrios Markou

The [NatureCounts_SpatialData_Tutorial](https://github.com/BirdsCanada/NatureCounts_SpatialData_Tutorial) modules aim to help users integrate NatureCounts data with spatial and environmental layers in R and meet their conservation and research goals. In **Module 1: Spatial Data Exploration**, you distinguished between spatial data types (vector and raster) and explored NatureCounts data using a spatio-temporal map.

Mapping the spatial distribution of avian and environmental data is critical to our understanding of ecological processes. Conservation actions are most effective when focused on areas of ecological importance. These areas typically host high levels of biodiversity, concentrations of rare species, or species at risk. In this way, analyzing NatureCounts data within specific boundaries might be relevant to your work. 

# 2.0 Learning Objectives 

By the end of **Module 2 - Conservation Areas**, users will know how to:

-   Read, preprocess, and visualize spatial vector data in R within areas of significant conservation potential: [Key Biodiversity Areas (KBAs)](https://kbacanada.org/about/) and [Priority Places for Species at Risk](https://environmental-maps.canada.ca/CWS_Storylines/index-ca-en.html#/en/priority_places-lieux_prioritaires).

-   Apply basic geoprocessing functions to spatial objects

The data used in this tutorial are downloaded from NatureCounts using R, the [KBA Canada Map Viewer](https://kbacanada.org/explore/map-viewer/), and the [Priority Places - Open Government Portal](https://open.canada.ca/data/en/dataset/91219d24-e877-4c8a-8bd2-b2b662e573e0).

This R tutorial requires the following **packages**:

```{r, message = FALSE}
library(naturecounts)
library(sf)
library(terra)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(ggspatial)
library(mapview)
library(leaflet)
library(leaflet.extras)
library(leaflet.providers)
```

*Example 2* You want to assess the spatial distribution of Wood Ducks (Ontario Breeding Bird Atlas data) across two conservation areas: KBAs across Ontario and the Long Point Walsingham Forest Priority Place.

KBAs: Navigate to the [KBA Canada Map Viewer](https://kbacanada.org/explore/map-viewer/), search for all the KBA's in Ontario, and download the .shp and .csv file to your working directory.

Priority Place: Downloaded from the [Priority Places - Open Government Portal](https://open.canada.ca/data/en/dataset/91219d24-e877-4c8a-8bd2-b2b662e573e0).

# 2.1 Key Biodiversity Areas

To view your working directory.

```{r}
getwd()
```

It helps if you assign these files a meaningful name! We can read in our KBA polygons using the `sf` package once it is in your working directory.

```{r}
ontario_kba <- sf::st_read("ontario_kba.shp")
```

`sf` objects are stored in R as a spatial dataframe which contains the attribute table of the vector along with the geometry type. When we examine the dataframe, it looks like there are many duplicate entries including duplicate geometries (vertices). To clean this up, we can apply the `st_make_valid()` and `distinct()` functions to our spatial dataframe:

```{r}
ontario_kba <- ontario_kba %>% st_make_valid() %>% distinct()
```

Our spatial data is also accompanied by a CSV file that contains additional useful attributes (landcover, species, etc) concerning our KBAs. Let's read in the accompanying CSV file for our KBA layer.

```{r}
kba_attributes <- read.csv("ontario_kba.csv")
```

Great! We can now join these dataframes using the handy `tidyverse` package. However, we'll want to select for specific columns first to avoid redundancies before performing our join:

```{r}
kba_attributes <- kba_attributes %>%
  select("SiteCode",
         "DateAssessed",
         "PercentProtected",
         "BoundaryGeneralized",
         "Level",
         "CriteriaMet",
         "ConservationActions",
         "Landcover",
         "Province",
         "Species")
```

Both dataframes now contain unique columns, after our selection. We apply the `full_join()` function to hold all attributes within one dataframe.

```{r}
ontario_kba <- full_join(ontario_kba, kba_attributes, by = "SiteCode")
```

We can visualize the **ontario_kba** data with an interactive map, using the `leaflet` package.

```{r}
leaflet(width = "100%") %>%
  addTiles() %>%
  addPolygons(data=ontario_kba,color = "black", weight = 2, smoothFactor = 1,opacity = 1.0, fillOpacity = 0.5, fillColor = "red") %>% addFullscreenControl() %>%
  addLegend(colors = c("red"),labels = c("Ontario KBAs"),position = "bottomright")
```

Similarly, the package `mapview` (based on leaflet) can also be used to make interactive plots. We can represent specific attributes like so:

```{r}
mapview::mapview(ontario_kba, zcol = "PercentProtected")
```

In this example, were interested in all the KBA polygons of Ontario. However, if you were working with a larger data set, it is possible to filter your dataframe to retrieve only specific polygons that meet certain criteria relevant to your research. To do so, we can apply filters based on a variable condition. For example, say we only wanted KBA's greater than 100km\^2 in size:

> To execute this code chunk, remove the \#

```{r}
# kba_name <- ontario_kba %>% 
  # dplyr::filter(Area > 100) 
```

Let's search NatureCounts for the Ontario Breeding Bird Atlas (ONATLAS3PC) dataset to assess Wood Duck distribution across Ontario KBA's. The steps here are similar to those you used in the Cedar Waxwing example.

```{r}
collections <- meta_collections()
View(meta_collections())
```

```{r}
search_species("wood duck")
```

Then let's filter and download data specifically for the Wood Ducks in Ontario.

```{r}
atlas_on <- nc_data_dl(collections = "ONATLAS3PC", species = 360, username = "testuser", info = "spatial_data_tutorial")
```

We can then convert our NatureCounts data into a spatial object. To do so, we deploy the `st_as_sf` function and specify the coordinate reference system (CRS).

The CRS of our KBA sf object can be returned with `st_crs()`.

```{r}
sf::st_crs(ontario_kba)
```

Our KBA sf object is stored with World Geodetic System 1984 (WGS 84) coordinates, EPSG = 4326. Now we can convert our **atlas_on** dataframe to an sf object using the same CRS.

```{r}
atlas_on_sf <- sf::st_as_sf(atlas_on,
                        coords = c("longitude", "latitude"), crs = 4326)
```

Now let's ensure that the conversion was successful. You'll notice a new geometry column where each observation is a point.

```{r}
str(atlas_on_sf) # view the sf object
```

There are a few other useful functions that may be applied to sf objects:

| `sf::st_transform()` - transforms the CRS of a specified CRS object
| `sf::st_drop_geometry()` - removes the geometry column of a sf object

For example, `st_transform()` can be applied to project our spatial object using a different CRS like NAD83 / UTM zone 16N (EPSG = 26916).

```{r}
ontario_kba <- sf::st_transform(ontario_kba, crs = 26916) 
```

It can also be used to ensure that the CRS of our spatial objects match.

```{r}
atlas_on_sf <- st_transform(atlas_on_sf, crs = st_crs(ontario_kba))
```

Geoprocessing functions allow us to manipulate or compute spatial objects based on interactions between their geometries. There are several useful functions integrated into the `sf` package including:

| `st_intersection(x, y)` - creates geometry of the shared portion of x and y
| `st_crop(x, y, ..., xmin, ymin, xmax, ymax)` - creates geometry of x that intersects a specified range
| `st_difference(x, y)` - creates geometry from x that does not intersect with y
| `st_area`, `st_length`, and `st_distance` can also be used to compute geometric measurements

More resources, including as `sf` package **cheatsheat** can be found [here](https://github.com/r-spatial/sf).

To identify Wood Ducks observed in Ontario KBA's, we can apply the `st_intersection()` function.

```{r}
wood_ducks_kba <- sf::st_intersection(ontario_kba, atlas_on_sf)
```

You will get a `warning` message, which you can safely ignore.

```{r}
# If need be, transform your spatial data back to EPSG:4326 to visualize with leaflet
ontario_kba <- st_transform(ontario_kba, crs = 4326)
wood_ducks_kba <- st_transform(wood_ducks_kba, crs = 4326)
```

Apply leaflet once more to visualize our polygon data and point data, this time using the `addCircleMarkers` argument.

```{r}
leaflet(width = "100%") %>%
  addTiles() %>%
  addPolygons(data = ontario_kba, color = "black", weight = 2, smoothFactor = 1, 
  opacity = 1.0, fillOpacity = 0.5, fillColor = "violet") %>%
  addCircleMarkers(data = wood_ducks_kba, radius = 3, color = "orange", 
                   stroke = FALSE, fillOpacity = 0.8) %>%
  addFullscreenControl() %>%
  addLegend(colors = c("violet", "orange"), labels = c("Ontario KBA", "Wood Duck Observations"), position = "bottomright")
```

After geoprocessing our data in R, we can write out any sf objects to Shapefiles on a disk, where the argument delete_layer = TRUE is used to overwrite a file.

```{r}
# st_write(wood_ducks_kba,"wood_ducks_kba.shp", driver = "ESRI Shapefile", delete_layer = TRUE)
```

# 2.2 Priority Places

First, lets create a path to our downloaded Priority Place file.

```{r}
gdb_path <- "PriorityPlaces.gdb"
```

Then, let's inspect the spatial data.

```{r}
gdb_layers <- st_layers(gdb_path)
print(gdb_layers)
```

To read in our spatial data object, we apply the `st_read` function and specify our desired data layer.

```{r}
priori_place_polygons <- st_read(dsn = gdb_path, layer = "PriorityPlacesBoundary")
```

Were interested in Wood Duck distribution across the Long Point Walsingham Forest Priority Place. We'll use `dplyr` to filter based on a variable condition.

```{r}
long_point_polygon <- priori_place_polygons %>%
  dplyr::filter(Name == "Long Point Walsingham Forest") # filters based on multipolygon name 
```

Then reproject the Wood Duck data to match our Priority Place using `st_transform`.

```{r}
atlas_on_sf <- st_transform(atlas_on_sf, crs = st_crs(long_point_polygon))
```

Next, we'll apply our geoprocessing function to find the Wood Duck observations that intersect with our chosen Priority Place.

```{r}
wood_ducks_longpoint <- sf::st_intersection(long_point_polygon, atlas_on_sf)
```

Finally, we'll transform our spatial objects one more time before visualizing them with `leaflet`.

```{r}
long_point_polygon <- st_transform(long_point_polygon, crs = 4326)
```

```{r}
wood_ducks_longpoint <- st_transform(wood_ducks_longpoint, crs = 4326)
```

```{r}
leaflet(width = "100%") %>%
  addTiles() %>%
  addPolygons(data = long_point_polygon, color = "black", weight = 2, smoothFactor = 1, 
  opacity = 1.0, fillOpacity = 0.5, fillColor = "red") %>%
  addCircleMarkers(data = wood_ducks_longpoint, radius = 5, color = "green", 
                   stroke = FALSE, fillOpacity = 0.8) %>%
  addFullscreenControl() %>%
  addLegend(colors = c("red", "green"), labels = c("Long Point Walsingham Forest", "Wood Duck Observations"), position = "bottomright")
```

**Congratulations**! You completed **Module 2 - Spatial Data Manipulation: KBA's and Priority Places**. Here, you successfully filtered and visualized NatureCounts data using spatio-temporal, KBA, and Priority Place maps. In [Module 3](03-ClimateData.Rmd), you can explore more spatial data visualization while linking climate data to NatureCounts observations.