---
title: "02-ManipSpatialData_Part2"
author: "Dimitrios Markou"
date: "2024-07-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# 2.? Organizing spatial data downloads

Spatial data can be complex and storage-intensive and become difficult to access

We also might want to rename our downloaded files to something more meaningful or distinguishable on our disk. To do so, we provide this helpful function:

```{r}
# Define the original base name and the new base name
original_base_name <- "old_shapefile_name"
new_base_name <- "new_shapefile_name"

# Define the directory containing the shapefile
shapefile_directory <- "path/to/shapefile/directory"

# List of shapefile extensions
shapefile_extensions <- c(".shp", ".shx", ".dbf", ".prj", ".cpg")

# Function to rename files
rename_shapefile <- function(original_base_name, new_base_name, shapefile_directory, extensions) {
  for (ext in extensions) {
    original_file <- file.path(shapefile_directory, paste0(original_base_name, ext))
    new_file <- file.path(shapefile_directory, paste0(new_base_name, ext))
    
    if (file.exists(original_file)) {
      file.rename(original_file, new_file)
    } else {
      message(paste("File not found:", original_file))
    }
  }
}

# Rename the shapefile and its accompanying files
rename_shapefile(original_base_name, new_base_name, shapefile_directory, shapefile_extensions)
```


