---
title: "03-EnvironmentalData"
author: "Dimitrios Markou"
date: "2024-08-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chapter 3 Environmental Data

The [Whooping Crane nesting area and summer range](https://kbacanada.org/site/?SiteCode=NT002) is located approximately 75 km west of Fort Smith, Northwest Territories.

```{r}
library(naturecounts)
library(sf)
library(terra)
library(luna)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(ggspatial)
library(mapview)
library(leaflet)
library(leaflet.extras)
library(leaflet.providers)
```

# Visualizing the study area 

Our first step is to get an appreciation for the study area by downloading the KBA boundary / environmental data and rendering a quick map.

Data downloads can sometimes have clunky names which are the opposite of R-friendly. Here is a useful function we can apply to our downloaded KBA shapefile and accompanying files to change their basename:

```{r}
rename_shapefile <- function(directory, old_name, new_name) {
  # Define the file extensions typically associated with shapefiles
  extensions <- c(".shp", ".shx", ".dbf", ".prj", ".sbn", ".sbx", ".cpg", ".xml", ".qix")
  
  # Loop through each extension and rename the corresponding file
  for (ext in extensions) {
    old_file <- file.path(directory, paste0(old_name, ext))
    new_file <- file.path(directory, paste0(new_name, ext))
    
    # Check if the old file exists before attempting to rename
    if (file.exists(old_file)) {
      file.rename(old_file, new_file)
    }
  }
}

  # Apply the function 
rename_shapefile("C:/Users/dimit/Birds Canada/NatureCounts_SpatialData_Tutorial/Data/KBA/WhoopingCrane_KBA", "kba.20240822070858", "whooping_crane_kba")
```

Read in the KBA boundary 

```{r}
whooping_crane_kba <- sf::st_read("Data/KBA/WhoopingCrane_KBA/whooping_crane_kba.shp")
```
```{r}
whooping_crane_kba <- whooping_crane_kba %>% st_make_valid() %>% distinct()
```

```{r}
kba_attributes <- read.csv("Data/KBA/WhoopingCrane_KBA/whooping_crane_kba.csv")
```

```{r}
kba_attributes <- kba_attributes %>%
  dplyr::select("SiteCode",
         "DateAssessed",
         "PercentProtected",
         "BoundaryGeneralized",
         "Level",
         "CriteriaMet",
         "ConservationActions",
         "Landcover",
         "Province",
         "Species")
```

```{r}
whooping_crane_kba <- full_join(whooping_crane_kba, kba_attributes, by = "SiteCode")
```

```{r}
leaflet(width = "100%") %>%
  addTiles() %>%
  addPolygons(data = whooping_crane_kba, color = "black", weight = 2, smoothFactor = 1,
              opacity = 1.0, fillOpacity = 0.6, fillColor = "limegreen") %>%
  addFullscreenControl() %>%
  addLegend(colors = c("limegreen"), 
            labels = c("Whooping Crane Nesting Area and Summer Range"), 
            position = "bottomright") %>%
  addScaleBar(position = "bottomleft", options = scaleBarOptions(imperial = FALSE))
```

# Retrieving Whooping Crane data from NatureCounts

```{r}
collections <- meta_collections()
View(meta_collections())
```

Upon inspection, we see that there are at least two ATLAS collections relevant to our research. The ABATLAS1 and ABATLAS2 collections both cover a wide range of survey years (1987-1991 and 2000-2005, respectively). 

```{r}
search_species("whooping crane")
```
Let's download data from all collections using the relevant project_id and species_id: 

```{r}
whooping_crane_data <- nc_data_dl(collections = c("ABATLAS1", "ABATLAS2"), species = 4030, username = "dimimarkou", info = "spatial_data_tutorial")
```
Convert the ATLAS data into a spatial object: 

```{r}
whooping_crane_data_sf<- sf::st_as_sf(whooping_crane_data,
                        coords = c("longitude", "latitude"), crs = 4326)
```

To write this spatial data object as a shapefile: 

```{r}
sf::st_write(whooping_crane_data_sf, "C:/Users/dimit/Birds Canada/NatureCounts_SpatialData_Tutorial/Data/Priority Places/WhoopingCrane/whooping_crane_data.shp")
```
```{r}
sf::st_write(whooping_crane_kba, "C:/Users/dimit/Birds Canada/NatureCounts_SpatialData_Tutorial/Data/Priority Places/WhoopingCrane/whooping_crane_kba.shp")
```



To apply geoprocessing function, we have to reproject our KBA and ATLAS spatial data layers:

```{r}
whooping_crane_kba <- sf::st_transform(whooping_crane_kba, crs = 26916) 

whooping_crane_data_sf <- st_transform(whooping_crane_data_sf, crs = st_crs(whooping_crane_kba))
```

Perform an intersection to find which observations are within the target KBA:

```{r}
whooping_crane_data_sf <- sf::st_intersection(whooping_crane_kba, whooping_crane_data_sf)
```

Transform both spatial data layers back to a CRS suitable for mapping:

```{r}
whooping_crane_kba <- st_transform(whooping_crane_kba, crs = 4326)
whooping_crane_data_sf <- st_transform(whooping_crane_data_sf, crs = 4326)
```

Visualize the Whooping Crane observations within the KBA:

```{r}
leaflet(width = "100%") %>%
  addTiles() %>%
  addPolygons(data = whooping_crane_kba, color = "black", weight = 2, smoothFactor = 1,
              opacity = 1.0, fillOpacity = 0.6, fillColor = "limegreen") %>%
   addCircleMarkers(data = whooping_crane_data_sf, radius = 3, color = "black", stroke = FALSE, fillOpacity = 0.8) %>%
  addFullscreenControl() %>%
  addLegend(colors = c("limegreen"), 
            labels = c("Whooping Crane Nesting Area and Summer Range"), 
            position = "bottomright") %>%
  addScaleBar(position = "bottomleft", options = scaleBarOptions(imperial = FALSE))
```

# Downloading MODIS data 

```{r}
modis_products <- getProducts("^MOD|^MYD|^MCD")
head(modis_products)
```
```{r}
product <- "MCD12Q1"
productInfo(product)
```
MCD12Q1 provides global land cover types at yearly intervals (2001-2022). We'll grab data from years corresponding to the ABATLAS2 surveys:

```{r}
start <- "2000-01-01"
end <- "2006-01-01"
```

```{r}
datadir <- file.path("C:/Users/dimit/Birds Canada/NatureCounts_SpatialData_Tutorial/Data", "MODIS")
dir.create(datadir, showWarnings = FALSE)
```

```{r}
modis_data <- luna::getNASA(product, start, end, aoi= whooping_crane_data_sf, download=TRUE,
                     path=datadir, username="dimimarkou", password=",xx4=_#Q?VPBxBm")
modis_data
```

```{r}
# List of HDF files
hdf_files <- list.files(path = "C:/Users/dimit/Birds Canada/NatureCounts_SpatialData_Tutorial/Data/MODIS", pattern = "\\.hdf$", full.names = TRUE)

length(hdf_files)
```
Inspect the data: 

```{r}
file_name <- basename(hdf_files)

date_ts <- str_extract(file_name, pattern = "[0-9]{7}")

date_ts <- str_replace(date_ts,
                       pattern = "A",
                       replacement = "")

date_ts <- lubridate::as_date(date_ts, format = "%Y%j")

date_data <- as.data.frame(date_ts)

year <- year(date_ts)

date_data <- cbind(date_data, year)

df_group <- group_by(date_data, year)

date_data <- summarize(df_group, n())

print(date_ts)

print(date_data)
```

```{r}
datadir <- "C:/Users/dimit/Birds Canada/NatureCounts_SpatialData_Tutorial/Data/MODIS"
mf <- file.path(datadir, "MCD12Q1.A2001001.h12v02.061.2022146050005.hdf")

library(terra)
r <- rast(mf[1])
r
```
```{r}
sds_list <- sds(mf)
print(sds_list)
```






