---
title: "03-EnvironmentalData"
author: "Dimitrios Markou"
date: "2024-08-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chapter 3 Environmental Data

##### Author: Dimitrios Markou

# 3.0 Learning Objectives

By the end of **Chapter 3 - EnvironmentalData**, users will know how to:

-   Process Digital Terrain Model (DTM) and Landcover data
-   Link NatureCounts data to environmental data for analysis
-   Calculate species diversity indices

This R tutorial requires the following packages:

```{r, eval = TRUE, warning = FALSE, message = FALSE}
library(naturecounts)
library(sf)
library(terra)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(ggspatial)
library(mapview)
library(leaflet)
library(leaflet.extras)
library(leaflet.providers)
```

This tutorial uses the following spatial data:

1.  [Places administered by Parks Canada](https://open.canada.ca/data/en/dataset/e1f0c975-f40c-4313-9be2-beb951e35f4e/resource/0744186a-449b-4f1f-9f1e-952a94c6d5ca) - Boundary shapefiles

2.  [Forêt ouverte (lidar-derived products)](https://www.donneesquebec.ca/recherche/dataset/produits-derives-de-base-du-lidar/resource/1fb86026-791c-45d2-88d7-58754ecebd73) - Digital Terrain Models (DTMs)

3.  [2015 Land Cover of Canada](https://open.canada.ca/data/en/dataset/4e615eae-b90c-420b-adee-2ca35896caf6) - Landcover map (30 m resolution)

# 3.1 Study Area - La Mauricie National Park

[La Mauricie National Park](https://parks.canada.ca/pn-np/qc/mauricie/nature) is situated in the Laurentian mountains and covers 536 km2 within the Eastern Canadian Temperate-Boreal Forest transition ecoregion. The environment is characterized by mixed forests, lakes, rivers, and hills that range from from 150 m to over 500 m in elevation. The park provides suitable habitat for a variety of wildlife including at least 215 bird species.

Read the National Park polygons into R.

```{r eval = FALSE}
national_parks <- st_read("path/to/your/shp")
```

```{r, message = FALSE, eval = TRUE, echo = FALSE}
national_parks <- st_read("C:/Users/dimit/Birds Canada/Data/nationalparks/vw_Places_Public_lieux_public_APCA.shp")
```

Filter the national_parks dataset for La Mauricie National Park.

```{r}
mauricie_boundary <- national_parks %>%
  filter(OBJECTID == "21")
```

Write the boundary as a shapefile to your disk.

```{r}
# Drop the Z-dimension (3D component) to make it 2D
mauricie_boundary_2D <- st_zm(mauricie_boundary, drop = TRUE, what = "ZM")

# Write the 2D mauricie_boundary object as a shapefile to your disk
st_write(mauricie_boundary_2D, "C:/Users/dimit/Birds Canada/Data/mauricie/mauricie_boundary.shp")
```
```{r}
st_write(mauricie_boundary_2D, "C:/Users/dimit/Birds Canada/Data/mauricie/mauricie_boundary.kml", driver = "KML")
```


# 3.2 NatureCounts Data

To assess the species distribution within the National Park, we will use data from the [Quebec Breeding Bird Atlas (2010 - 2014)](https://naturecounts.ca/nc/default/datasets.jsp?code=QCATLAS2PC&sec=bmdr) which is part of a 5 year project that covers the distribution and abundance of all species breeding in the province.

Download NatureCounts data:

```{r}
quebec_atlas <- nc_data_dl(collections = "QCATLAS2PC", username = "testuser", info = "spatial_data_tutorial")
```

eBird has the greatest number of provincial bird records, however, this collection comprise data of Access Level 4. If you wish to access this collection you must sign up for a free account and make a data request. Otherwise, you can carry forward with the tutorial without these data and skip this code chunk.

```{r, eval = FALSE}
ebird_quebec <- nc_data_dl(collections = c("QCATLAS2PC", "EBIRD-CA-QC"), username = "dimimarkou", info = "spatial_data_tutorial")
```

To create date and doy columns and ensure that the ObservationCount column is in the correct numeric format we can apply the `format_dates()` and `mutate()` functions. We will also filter the dataset to exclude rows with missing coordinates.

```{r}
quebec_atlas <- quebec_atlas %>%
  format_dates() %>%  # create the date and doy columns 
  mutate(ObservationCount = as.numeric(ObservationCount)) %>%  # convert to numeric format
  filter(!is.na(longitude) & !is.na(latitude))  # remove rows with missing coordinates
```

To convert the NatureCounts data to a spatial object and transform its crs to match the National Park boundary we can use the `st_as_sf()` and `st_transform()` functions, respectively.

```{r}
quebec_atlas_sf <- sf::st_as_sf(quebec_atlas,
                        coords = c("longitude", "latitude"), crs = 4326) # converts the quebec_atlas data to an sf object

mauricie_boundary <- st_transform(mauricie_boundary, crs = st_crs(quebec_atlas_sf)) # match the CRS
```

Clip the NatureCounts data to the National Park boundary using `st_intersection()`.

```{r}
mauricie_birds <- sf::st_intersection(quebec_atlas_sf, mauricie_boundary)
```

Tidyverse functions can help us summarize our data in a variety of ways. For example, if we wanted to determine the annual bird count for each year across all sites, we could use `mutate()` and `lubridate()` to extract the survey **year** from the **date** column. The `group_by()` function can then be used to group the observations by year, and `summarise()` can help calculate and create the **annual_count** column. Here, we ensure that the **ObservationCount** is in the correct format by applying `as.numeric()`.

```{r}
mauricie_birds_summary <- mauricie_birds %>%
  mutate(year = lubridate::year(date)) %>% # extracts the survey year using lubridate
  group_by(year) %>%
  summarise(annual_count = sum(as.numeric(ObservationCount), na.rm = TRUE)) %>% # calculates the annual_count
  filter(!is.na(year))  # remove rows with missing year

mauricie_birds_summary
```

If you wanted to summarize total count for each species at each site, respectively you could adjust the pipe like so using `pivot_wider()`.

```{r}
mauricie_species_summary <- mauricie_birds %>%
  st_drop_geometry() %>%  # drop the geometry column
  group_by(species_id, SiteCode) %>%
  summarise(total_count = sum(as.numeric(ObservationCount), na.rm = TRUE)) %>% # calculates the total_count column
  pivot_wider(names_from = species_id, # populates the column names with each species_id code
              values_from = total_count, # populates each cell with total_count
              values_fill = list(total_count = 0)) %>% # missing values are zero-filled
  group_by(SiteCode)

mauricie_species_summary
```

# 3.3 DEM, DSM & DTM

Digital elevation datasets store topographic information like elevation or slope and are a useful way to interpret landscape ecology. These datasets, i.e. Digital Elevation Models (DEMs), Digital Surface Models (DSMs), and Digital Terrain Models (DTMs) are derived through a variety of remote sensing and spatial interpolation techniques and all help describe land features.

**Digital Elevation Model** - represents the bare-Earth surface and [excludes]{.underline} all terrain vector features (i.e. streams, breaklines, and ridges), and all ground objects (power lines, buildings, trees, and vegetation).

**Digital Surface Model** - represents the heights of the Earth's surface and [includes]{.underline} all natural and artificial features or ground objects.

**Digital Terrain Model** - represents the bare-Earth surface topography and includes all terrain vector features. It does not include natural or artificial ground objects. In other words, it is a DEM that is augmented by the presence of streams, breaklines, and ridges.

Light Detection and Ranging (LiDAR) is an active remote sensing technology. It is performed using laser scanners that emit pulses of light and determine the position of target 3D objects by measuring the amount of time between pulses being emitted and received. It is a revolutionary technology that helps in the acquisition of extremely accurate land data over wide spatial and temporal scales.

Set the path to your TIF file directory.

```{r, eval = FALSE}
dir_path <- "path/to/your/directory"
```

```{r, eval = TRUE, echo = FALSE}
dir_path <- "C:/Users/dimit/Birds Canada/Data/mauricie"
```

Create a mosaic of the adjacent DTM rasters.

```{r, eval = TRUE}
# list all the TIFF files in your directory
dtm_files <- list.files(dir_path, pattern = "\\.tif$", full.names = TRUE)

# Read into a list of SpatRaster objects
dtm_list <- lapply(dtm_files, rast)

# Combine into a single SpatRaster object
dtm_mosaic <- do.call(mosaic, dtm_list)

# Print information about the stack
print(dtm_mosaic)
```

Let's check if the DTM and National Park boundary have the same crs by using the `st_crs()` function and equality operator (**==**) which will generate either TRUE or FALSE.

```{r, eval = TRUE}
st_crs(dtm_mosaic) == st_crs(mauricie_boundary)
```

To reproject the spatial data with the same CRS, we can use the `st_transform()` function.

```{r}
mauricie_boundary <- st_transform(mauricie_boundary, crs = st_crs(dtm_mosaic))
```

We can then use the base `plot()` function to visualize the terrain and boundary of La Mauricie National Park.

```{r regional_dtm, eval = TRUE, warning = FALSE}
# Plot the DTM mosaic
plot(dtm_mosaic, main = "DTM Mosaic with La Mauricie Boundary")

# Overlay the National Park boundary
plot(mauricie_boundary, add = TRUE, border = "red", col = NA, lwd = 2)
```

# 3.4 Cropping and masking

**Cropping** reduces the extent of a raster to the extent of another raster or vector.

**Masking** assigns NA values to cells of a raster not covered by a vector.

To mask a raster to a vector extent we can apply the `mask()` function from the `terra` package which uses the SpatVector format. Here, we create the mask while converting the National Park Boundary to a SpatVector using `vect()`.

```{r, warning = FALSE}
mauricie_mask <- mask(dtm_mosaic, vect(mauricie_boundary))
```

Visualize the regional and masked DTM's with a two-panel plot.

```{r masked_dem, eval = TRUE, warning = FALSE}
# Set up a 1x2 plotting layout
par(mfrow = c(1, 2))

# Plot dtm_mosaic
plot(dtm_mosaic, main = "regional DTM")

# Overlay the National Park boundary on the first plot
plot(mauricie_boundary, add = TRUE, border = "red", col = NA, lwd = 2)

# Plot mauricie_mask
plot(mauricie_mask, main = "masked DTM")

# Overlay the National Park boundary on the second plot
plot(mauricie_boundary, add = TRUE, border = "red", col = NA, lwd = 2)
```

# 3.5 Reclassification

When communicating our data, it might be useful to reclassify a range of raster values into classes. Reclassifying rasters is easily achieved using `terra` and the `classify()` function.

First, provide a reclassification table in the form of a matrix which specifies the range of raster values (columns 1 and 2) and the class ID (column 3) like so:

```{r}
dtm_rcl_matrix <- matrix(c(0, 200, 1,
                           200, 400, 2,
                           400, 1000, 3), byrow = TRUE, ncol = 3)
```

Then, apply the `classify()` function using the reclassification matrix:

```{r}
dtm_mosaic_reclass <- classify(dtm_mosaic, dtm_rcl_matrix)
```

Finally, give meaningful labels to each of the classes:

```{r, eval = TRUE}
elevation_levels <- data.frame(ID = c(1L, 2L, 3L),
                               category = c("Low elevation",
                                            "Mid elevation",
                                            "High elevation"))
levels(dtm_mosaic_reclass) <- elevation_levels
```

Visualize the regional DTM and National Park boundary using the reclassification scheme:

```{r, eval = TRUE, warning = FALSE}
# Set up a 1x2 plotting layout
par(mfrow = c(1, 2))  # Set up a 1-row, 2-column plotting layout

# Plot the original DTM mosaic
plot(dtm_mosaic, main = "Regional DTM")

# Overlay the Mauricie boundary on the first plot
plot(mauricie_boundary, add = TRUE, border = "red", col = NA, lwd = 2)

# Plot the reclassified DTM mosaic
plot(dtm_mosaic_reclass, main = "Reclassified DTM")

# Overlay the Mauricie boundary on the second plot
plot(mauricie_boundary, add = TRUE, border = "red", col = NA, lwd = 2)

# Reset the plotting layout to the default
par(mfrow = c(1, 1))

```

# 3.6 Summary Statistics

To calculate the summary statistics for the values of individual SpatRaster layers you can use the `global()` function like so:

```{r}
global(mauricie_mask, fun = "mean", na.rm = TRUE)
```

It is also possible to make and implement a function to calculate multiple summary statistics simultaneously. For example, `my_summary_stats` calculates the max, min, mean and sd of an input numeric vector `x` and returns the summary statistics in a vector (give this a few minutes):

```{r}
my_summary_stats <- function(x) {
  out <- c(mean = mean(x, na.rm = TRUE), 
              sd = sd(x, na.rm = TRUE), 
              min = min(x, na.rm = TRUE), 
              max = max(x, na.rm = TRUE))
  return(out)
}
global(mauricie_mask, fun = my_summary_stats)
```

You can also create a histogram of the raster DTM data.

```{r, warning = FALSE}
terra::hist(mauricie_mask, axes = TRUE, 
            main = "Elevation values within La Mauricie National Park",
            xlab = "Elevation (m)",
            col = "Green")
```

# 3.7 Mapping

```{r}
# Group by SiteCode and summarize total_count
mauricie_site_summary <- mauricie_birds %>%
  group_by(SiteCode) %>%
  summarize(total_count = sum(ObservationCount, na.rm = TRUE))

mauricie_site_summary <- st_transform(mauricie_site_summary, crs = st_crs(mauricie_mask)) # Match the CRS

# Plot the mauricie_mask raster
plot(mauricie_mask, main = "Mauricie Mask with Site Summary Points")

# Overlay mauricie_site_summary multipoints
plot(st_geometry(mauricie_site_summary), 
     add = TRUE, 
     pch = 19,       # Solid circle
     col = "blue",   # Point color
     cex = 0.5)      # Point size
```

To extract the elevation values at each site.

```{r}
mauricie_birds <- st_transform(mauricie_birds, crs(mauricie_mask)) # Match the CRS

mauricie_birds_vect <- vect(mauricie_birds) # Converts to SpatVector format

elevation_values <- terra::extract(mauricie_mask, mauricie_birds_vect) # Extracts the elevation values for each point

elevation_values <- cbind(record_id = mauricie_birds$record_id, elevation = elevation_values[, 2]) # Creates the record_id and elevation columns 
```

To include the elevation values to the mauricie_birds dataset.

```{r}
mauricie_birds$elevation <- elevation_values[, 2] # attaches the 2nd column of the elevation_values dataset to mauricie_birds
```

# 3.8 Landcover Data

**Land use** describes the purpose that the land serves i.e. recreation, wildlife habitat, agriculture, etc.

**Land cover** describes the surface cover on the ground i.e. vegetation, urban, water, bare soil, etc.

See [Land Cover & Land Use](https://natural-resources.canada.ca/maps-tools-and-publications/satellite-imagery-elevation-data-and-air-photos/tutorial-fundamentals-remote-sensing/educational-resources-applications/land-cover-land-use/land-cover-land-use/9373) from Natural Resources Canada.

Read the landcover dataset using `terra::rast()`.

```{r}
landcover <- rast("C:/Users/dimit/Birds Canada/Data/landcover-2015-classification.tif")
landcover
```

Transform the crs of the National Park boundary to match that of the landcover dataset.

```{r}
mauricie_boundary <- st_transform(mauricie_boundary, crs = st_crs(landcover))
```

Crop the national dataset to reduce the extent of the raster to the extent of the Natiional Park.

```{r}
landcover_crop <- crop(landcover, vect(mauricie_boundary))
```

Plot the landcover raster and National Park boundary.

```{r, eval = TRUE, warning = FALSE}
plot(landcover_crop, axes = TRUE, plg = list(title = "Landcover"), main = "La Mauricie National Park - Landcover 2015")

# Overlay the Mauricie boundary on the first plot
plot(mauricie_boundary, add = TRUE, border = "red", col = NA, lwd = 2)
```

To extract all the pixel values found in the cropped raster.

```{r}
unique_vales <- unique(landcover_crop)
unique_vales
```

To determine the number of pixels associated with each pixel value.

```{r}
freq(landcover_crop)
```

# 3.9 Calculating NDVI using SENTINEL2 imagery

Describe the satellite image dataset

Summary
Name:
S2A_MSIL2A_20200619T154911_N0500_R054_T18TXS_20230428T011527.SAFE
Size:
1090MB
Sensing time:
2020-06-19T15:49:11.024000Z
Platform short name:
SENTINEL-2
Instrument short name:
MSI

Describe the relationship between bird presence and NDVI. 

In this section, you will read in Sentinel-2 satellite imagery over the region spanning La Mauricie National Park and calculate NDVI.


Read each band stored in the downloaded satellite imagery folder in your directory.

```{r}
bands <- list.files(path = "C:/Users/dimit/Birds Canada/Data/SENTINEL2/S2A_MSIL2A_20200619T154911_N0500_R054_T18TXS_20230428T011527.SAFE/GRANULE/L2A_T18TXS_A026077_20200619T155414/IMG_DATA/R10m", 
                    pattern = "\\.jp2$", full.names = TRUE)

# Read all the bands into a raster stack
sentinel_imagery <- rast(bands)

# Assign meaningful names to the bands
names(sentinel_imagery) <- c("AOT", "blue", "green", "red", "nir", "TCI_r", "TCI_g", "TCI_b", "WVP")

# Check the assigned names
print(names(sentinel_imagery))

# Check resolution, number of layers, and extent
res(sentinel_imagery)
nlyr(sentinel_imagery)
ext(sentinel_imagery)
```
Plot a true color composite of `sentinel_imagery` with a linear stretch.

```{r, eval = TRUE, warning = FALSE}
# Match the CRS of the raster stack and National Park boundary 
mauricie_boundary <- st_transform(mauricie_boundary, crs = crs(mauricie_sent))

# Plot the RGB composite
plotRGB(sentinel_imagery, r = 4, g = 3, b = 2, stretch = "lin", main = "RGB Composite with Boundary")

# Add the boundary shapefile to the plot
plot(mauricie_boundary, add = TRUE, border = "red", col = NA, lwd = 2)
```

Crop `sentinel_imagery` to the extent of `mauricie_boundary` then plot the cropped image in true color.

```{r}
mauricie_sl <- crop(sentinel_imagery, vect(mauricie_boundary))
terra::plotRGB(mauricie_sl, r = 4, g = 3, b = 2, stretch = "lin", axes = FALSE)
plot(st_geometry(mauricie_boundary), col = NA, border ="red", add = TRUE)
```

To avoid the effect of water when analyzing spectral indices like NDVI we can apply a water mask. 

The normalized difference water index (NDWI; [McFeeters,
1996](https://www.tandfonline.com/doi/abs/10.1080/01431169608948714)) is
a spectral index that can be used to detect water bodies from
multispectral imagery.

$$NDWI = \frac{GREEN - NIR}{GREEN + NIR}$$

```{r}
calc_ndwi <- function(green, nir) {
  ndwi <- c((green - nir)/(green + nir))
  return(ndwi)
}

# Calculate NDWI
mauricie_sl_ndwi <- calc_ndwi(mauricie_sl$green, mauricie_sl$nir)

# Create water mask
water_mask <- mauricie_sl_ndwi >= 0

# Set up the plot area for 1 row and 2 columns
par(mfrow = c(1, 2))

# Plot NDWI
plot(mauricie_sl_ndwi, main = "NDWI")

# Plot water mask 
plot(water_mask, main = "Water mask")

# Reset plotting layout to default
par(mfrow = c(1, 1))
```

Use the function `focal()` to apply a convolution filter returning the **mode** of each pixel of `water_mask` in a 3 x 3 windows of equal weights (argument `w = 3`). Name the output raster `water_mask_mode`. The *mode* of a vector `x` is the value that appears the most often in `x`.

```{r}
get_mode <- function(x, na.rm = TRUE) {
  
  if (na.rm) {
    x <- x[!is.na(x)]
  }
  
  ux <- unique(x)
  
  ux[which.max(tabulate(match(x, ux)))]
  
}
```

Calculate `water_mask_mode`.

```{r}
water_mask_mode <- focal(water_mask, w = 3, fun = get_mode)
```

Combine the `water_mask` and `water_mask_mode`. 

```{r}
water_mask_combined <- c(water_mask, water_mask_mode)
```

Assign meaningful names to each raster mask and plot them.

```{r, warning = False}
names(water_mask_combined) <- c("water mask", "water mask mode")

plot(water_mask_combined)
```
Apply the water mask to `mauricie_sl`.

```{r}
mauricie_sl_water_mask <- mask(mauricie_sl, water_mask_mode, maskvalues = 1)
```
Apply another mask to assign NA values to those pixels outside of the National Park boundary.

```{r}
mauricie_sl_mask <- mask(mauricie_sl_water_mask, vect(mauricie_boundary))
```
Plot a true color composite of the National Park with both the water and boundary masks applied.

```{r}
terra::plotRGB(mauricie_sl_mask, r = 4, g = 3, b = 2, stretch = "lin", axes = FALSE)
plot(st_geometry(mauricie_boundary), col = NA, border ="red", add = TRUE)
```
To calculate NDVI, create and apply the following function:

- $$NDVI = \frac{NIR - RED}{NIR + RED}$$ 

```{r}
calc_ndvi <- function(nir, red) {
  ndvi <- c((nir - red)/(nir  + red))
  return(ndvi)
}

# Calculate NDVI 
mauricie_sl_ndvi <- calc_ndvi(mauricie_sl_mask$nir, mauricie_sl_mask$red)
```

Plot NDVI and the National Park boundary.

```{r}
plot(mauricie_sl_ndvi, main = "NDVI values across La Mauricie National Park")
plot(st_geometry(mauricie_boundary), col = NA, border ="red", add = TRUE)
```
Extract NDVI values for each bird observation site and append it to `mauricie_birds`.


