---
title: "05-LandcoverData"
author: "Dimitrios Markou"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chapter 5: Land Cover Data

##### Authors: Dimitrios Markou, Danielle Ethier

> In [Chapter 4](04-ElevationData.Rmd), you processed Digital Terrain Models, applied crop and mask procedures, and extracted elevation values to combine with NatureCounts data. This chapter will build on these skills by extracting land cover data over [La Mauricie National Park](https://parks.canada.ca/pn-np/qc/mauricie/nature) and combine these values to NatureCounts data from the [Quebec Breeding Bird Atlas (2010 - 2014)](https://naturecounts.ca/nc/default/datasets.jsp?code=QCATLAS2PC&sec=bmdr).

**Land cover** describes the surface cover on the ground like vegetation, urban, water, and bare soil while **land use** describes the purpose that the land serves like recreation, wildlife habitat, and agriculture. See [Land Cover & Land Use](https://natural-resources.canada.ca/maps-tools-and-publications/satellite-imagery-elevation-data-and-air-photos/tutorial-fundamentals-remote-sensing/educational-resources-applications/land-cover-land-use/land-cover-land-use/9373) from Natural Resources Canada for more information.

**To proceed with this chapter, complete section 4.1 Data Setup from [Chapter 4: Elevation Data](04-ElevationData.Rmd)**. **This Chapter will guide you through how to download the National Park boundary and NatureCounts data which are used in this lesson. Alternatively, you can access these files in the [Git repository data folder](https://github.com/BirdsCanada/NatureCounts_SpatialData_Tutorial/tree/d83eed714373048c8f9fdb4ea4a17f1fd4bef3f7/Data) or the [Google Drive data folder](https://drive.google.com/drive/folders/1gLUC6fROl4kNBvTGselhZif-arPexZbY?usp=sharing).**

# 5.0 Learning Objectives {#5.0LearningObjectives}

By the end of **Chapter 5 - Land Cover Data**, users will know how to:

-   Load and interpret raster data color maps: [Raster Color Maps](#5.2ColorMaps)
-   Create buffers around observation sites (points): [Point Buffers](#5.3PointBuffers)
-   Calculate landscape metrics, including PLAND, NP, and ED: [Landscape Metrics](#5.4LandscapeMetrics)
-   Combine NatureCounts data with land cover data for analysis: [Map & Extract Land Cover Data](#5.5Map-Extract)

Load the required packages:

```{r, eval = TRUE, warning = FALSE, message = FALSE}
library(naturecounts)
library(sf)
library(terra)
library(tidyverse)
library(lubridate)
library(readr)
library(leaflet)
library(leaflet.extras)
library(landscapemetrics)
library(units)
```

This tutorial uses the following spatial data:

1.  [Places administered by Parks Canada](https://open.canada.ca/data/en/dataset/e1f0c975-f40c-4313-9be2-beb951e35f4e/resource/0744186a-449b-4f1f-9f1e-952a94c6d5ca) - Boundary shapefiles

2.  [Quebec Breeding Bird Atlas (2010 - 2014)](https://naturecounts.ca/nc/default/datasets.jsp?code=QCATLAS2PC&sec=bmdr) - NatureCounts bird observations

    These data are available for download via Section 4.1: Data Setup in [Chapter 4](04-ElevationData.Rmd), the [Git repository data folder](https://github.com/BirdsCanada/NatureCounts_SpatialData_Tutorial/tree/d83eed714373048c8f9fdb4ea4a17f1fd4bef3f7/Data) or the [Google Drive data folder](https://drive.google.com/drive/folders/1gLUC6fROl4kNBvTGselhZif-arPexZbY?usp=sharing)

3.  [2015 Land Cover of Canada](https://open.canada.ca/data/en/dataset/4e615eae-b90c-420b-adee-2ca35896caf6) - Land Cover map (30 m resolution)

# 5.1 Data Setup {#5.1DataSetup}

Read in the National Park boundary you saved or downloaded from [Chapter 4: Elevation Data](04-ElevationData.Rmd).

```{r}
mauricie_boundary <- st_read("mauricie_boundary.shp")
```

To read in the land cover dataset, navigate to [2015 Land Cover of Canada](https://open.canada.ca/data/en/dataset/4e615eae-b90c-420b-adee-2ca35896caf6), scroll down to **Data and Resources** and select the TIFF file to download. Extract the data download and save the file to your directory before applying `terra::rast()`.

```{r}
# if the data are in your working directory
landcover <- rast("landcover-2015-classification.tif")

# else, specify the location of your data
# landcover <- rast(path/to/your/landcover-2015-classification.tif)

print(landcover)
```

Transform the coordinate reference system (crs) of the National Park boundary to match that of the land cover dataset.

```{r}
mauricie_boundary <- st_transform(mauricie_boundary, crs = st_crs(landcover))
```

# 5.2 Spatial Extents: Crop & Mask

**Cropping** reduces the extent of a raster to the extent of another raster or vector.

To crop a raster we can apply the `crop()` function from the `terra` package which uses the SpatVector format. Here, we crop the extent of the landcoover raster to the extent of the National Park. To do this we need to convert the National Park Boundary to a SpatVector using `vect()`.

```{r}
landcover_crop <- crop(landcover, vect(mauricie_boundary))
print(landcover_crop)
```

**Masking** assigns NA values to cells of a raster not covered by a vector.

To mask the landcover raster to the National Parks vector extent we can apply the `mask()` function from the `terra` package which also uses the SpatVector format. This will apply NA values to all cells outside the extent of the National Park.

```{r}
landcover_mask <- mask(landcover_crop, vect(mauricie_boundary))
print(landcover_mask)
```

Visualize the cropped and masked land cover rasters with a two-panel plot.

```{r}
# Set up the plot area for 1 row and 2 columns
par(mfrow = c(1, 2))

# Plot the cropped land cover raster
plot(landcover_crop, main = "crop")

# Add National Park boundary
plot(st_geometry(mauricie_boundary), col = NA, border ="purple", lwd = 2, add = TRUE)

# Plot the masked land cover raster 
plot(landcover_mask, main = "mask")

# Add National Park boundary
plot(st_geometry(mauricie_boundary), col = NA, border ="purple", lwd = 2, add = TRUE)

# Reset plotting layout to default
par(mfrow = c(1, 1))
```

OPTIONAL: To write the masked raster to your disk, you can use the writeRaster() function from terra. **This raster will be required to complete [Chapter 7: Raster Summary Tools](07-RasterSummaryTools.Rmd)**. We have uploaded this file to the [**Google Drive data folder**](https://drive.google.com/drive/folders/1gLUC6fROl4kNBvTGselhZif-arPexZbY?usp=sharing) for your convenience in the subfolder labeled **mauricie**. 

To execute this code chunk, remove the \# and make sure to specify the file format (.tif) in the output path

```{r}
# output_path <- getwd()

# writeRaster(landcover_mask, output_path, overwrite = TRUE)
```

# 5.2 Raster Color Maps {#5.2ColorMaps} 

Color maps are attribute tables that associate a set of pixel values in categorical raster objects with specific colors. The color table is a data frame with up to four columns - red, green, blue, and alpha (opacity) associated to one value. These tables are useful for mapping and can be viewed or set using the `coltab()` function (see [here](https://rspatial.github.io/terra/reference/colors.html)). Saving a raster and its color table to a file will also preserve its color information.

Examine the contents of the masked land cover raster. The `freq()` function describes the pixel count per land cover class represented in the region.

```{r}
freq(landcover_mask)
```

View the color table for the masked land cover raster using the `coltab()` function. Save it to your environment as a dataframe called **landcover_rgb** using the `as.data.frame()` function.

```{r}
landcover_rgb <- as.data.frame(coltab(landcover_mask))
landcover_rgb
```

Filter the color table to only include those classes represented in the masked land cover raster. To assign the appropriate labels to each land cover class, refer to the [Class Index](https://open.canada.ca/data/en/dataset/ee1580ab-a23d-4f86-a09b-79763677eb47/resource/b8411562-49b7-4cf6-ac61-dbb893b182cc) and use the `mutate()` function after filtering for the relevant classes and tidying-up the attribute names.

```{r}
landcover_rgb <- landcover_rgb %>%
  rename_with(tolower) %>% # lowercase is more R friendly!
  rename(landcover_class = value) %>%
  mutate(landcover_class = as.factor(landcover_class)) %>%
  filter(landcover_class %in% c(1, 5, 6, 8, 10, 14, 16, 17, 18)) %>% # filter based on the classes represented in the "landcover_crop"
  mutate(landcover = c("needleleaf", "broadleaf", "mixed forest", "shrubland", "grassland", "wetland", "barren", "urban", "water")) # assign land cover class labels
```

Let's combine the color table with the raster data.

First, convert the cropped landcover raster to a dataframe.

```{r}
landcover_mask_df <- as.data.frame(landcover_mask, xy = TRUE, cells = TRUE)
```

Next, merge the two dataframes by performing a `left_join()`. Ensure that the grouping variable for both dataframe matches.

```{r}
landcover_mask_df <- landcover_mask_df %>%
  rename(landcover_class = Class) %>% 
  left_join(landcover_rgb, by = "landcover_class") 
```

Now lets map the cropped land cover raster using `ggplot()`.

```{r}
ggplot() +
  geom_raster(data = landcover_mask_df, aes(x = x, y = y, fill = landcover)) +
  scale_fill_manual(
    values = setNames(
      rgb(landcover_rgb$red, landcover_rgb$green, landcover_rgb$blue, maxColorValue = 255),
      landcover_rgb$landcover
    ),
    name = "Landcover"
  ) +
  labs(
    title = "La Mauricie National Park - Landcover 2015",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_bw() +
  coord_fixed() +
  geom_sf(data = mauricie_boundary, fill = NA, color = "purple", size = 1.5)
```

# 5.3 Point Buffers {#5.3PointBuffers}

Buffers allow us to generate polygons at a specified distance around points, lines or other polygons. Because birds use their habitat at the landscape-level and not at a single observation point, buffers allow us to summarize land cover and habitat information within the local neighborhood.

Read in the NatureCounts data you saved or downloaded from [Chapter 4: Elevation Data](04-ElevationData.Rmd).

```{r}
mauricie_birds_df <- read.csv("mauricie_birds_df.csv")
```

Create an `sf` object from the NatureCounts data that represents the unique point count locations. Because our land cover data is from 2015, we filter the NatureCounts data for the most recent survey year (2014) and for relevant attribute columns.

```{r}
mauricie_birds_local <- mauricie_birds_df %>% filter(survey_year == 2014) %>% 
  # distinct(longitude, latitude) %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  select(record_id, 
         species_id, 
         SiteCode, 
         Locality, 
         SamplingEventIdentifier, 
         RouteIdentifier, 
         english_name,
         ObservationCount,
         date,
         geometry)
```

Assign a unique identifier to each observation location.

```{r}
mauricie_birds_local <- mauricie_birds_local %>%
  mutate(point_id = paste0("point_id_", seq_len(nrow(.)))) %>%
  select(point_id, everything())
```

Use the `st_buffer` function to generate 3 km diameter circular neighborhoods centered at each observation location.

```{r}
mauricie_birds_buffer <- mauricie_birds_local %>%
  st_buffer(dist = set_units(1.5, "km"))
```

# 5.4 Landscape Metrics {#5.4LandscapeMetrics}

In landscape ecology, the composition and configuration of land types tells us what habitat is available and how its distributed spatially. **Landscape metrics** help us characterize these spatial patterns across landscapes or within specific areas like point buffers.

The `landscapemetrics` package can be used to calculate some common metrics that quantify habitat composition and configuration:

-   Percent landcover (PLAND) - percent of landscape of a given class

-   Edge density (ED) - total boundary length of all patches of a given class per unit area

-   Number of patches (NP) - number of unique contiguous units

-   Largest patch index (LPI) - percent of the landscape comprised of the single largest patch

-   Mean core area index (CAI_MN) - percent of the patch that is comprised of core area which is a compound measure of shape, area, and edge depth.

-   Patch Cohesion Index (COHESION) - area-weighted mean perimeter-area ratio that helps assess connectivity

> A "patch" is an intuitive concept that describes a contiguous group of cells of the same landcover category. Patches can be defined using the 4-neighbour or 8-neighbour rules.

For each point buffer, crop and mask the land cover data and calculate the class-level PLAND and ED within each point buffer. Store the result in a dataframe.

```{r}
lsm <- list()

for (i in seq_len(nrow(mauricie_birds_buffer))) {
  buffer_i <- st_transform(mauricie_birds_buffer[i, ], crs = crs(landcover))  # Match the CRS between the buffer zones and land cover
  
  # Crop and mask the landcover data with the buffer areas
  lsm[[i]] <- crop(landcover, buffer_i) %>% 
    mask(buffer_i) %>%  
    # Calculate landscape metrics
    calculate_lsm(level = "class", metric = c("pland", "ed")) %>% 
    # Add identifying variables
    mutate(point_id= buffer_i$point_id, 
           survey_year = buffer_i$survey_year) %>%  
    select(point_id, level, class, metric, value)
}

# Combine results into a single dataframe
lsm <- bind_rows(lsm)
```

Add land cover class labels to the dataframe containing the calculated landscape metrics. 

```{r}
# Define the land cover class labels 
lc_classes <- tibble(
  class = c(1, 5, 6, 8, 10, 14, 16, 17, 18),
  landcover = c("needleleaf", "broadleaf", "mixed forest", "shrubland", "grassland", "wetland", "barren", "urban", "water")
)

lsm <- inner_join(lsm, lc_classes, by = "class")
```

Calculate the mean class-level PLAND and ED across all point buffers.

```{r}
lsm_mean <- lsm %>%
  group_by(landcover, metric) %>% 
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop") 

# View the resulting dataframe
print(lsm_mean)
```

Compare the PLAND and ED values across each landcover class.

```{r}
ggplot(lsm_mean, aes(x = landcover, y = mean_value, color = metric)) +
  geom_point(size = 3, alpha = 0.8) +       # Add points
  labs(
    title = "Landscape Metrics (PLAND and ED) Across Landcover Classes",
    x = "Landcover Class",
    y = "Metric Value",
    color = "Metric"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels for clarity
    panel.grid.major.x = element_blank()   # Remove vertical gridlines 
  )
```

**Note**: Landscape pattern analysis requires careful consideration for the classification scheme, spatial extent, and choice of landscape metric - all of which should depend on the objectives of your study. How you group values also matters as it will influence how you interpret results.

# 5.5 Map & Extract Land Cover Data {#5.5Map-Extract}

Create a spatial dataframe from the NatureCounts data and join the unique point_id from the buffer geometry to the NatureCounts data by latitude and longitude using `st_join`.

```{r}
mauricie_birds_sf <- sf::st_as_sf(mauricie_birds_df, 
                        coords = c("longitude", "latitude"), crs = 4326)              # convert to sf object

mauricie_birds_sf <- st_join(mauricie_birds_sf, mauricie_birds_local, join = st_intersects)
```

Transform the CRS of the National Park boundary to match the NatureCounts data.

```{r}
mauricie_boundary <- st_transform(mauricie_boundary, crs = st_crs(mauricie_birds_sf)) # match the CRS
```

Summarize the NatureCounts data by point_id for mapping.

```{r}
# Group by SiteCode and summarize total_count
mauricie_site_summary <- mauricie_birds_sf %>%
  group_by(point_id) %>% filter(!is.na(point_id)) %>% # Filter out NA values
  summarize(total_count = sum(ObservationCount, na.rm = TRUE)) 
```

Next we can map both the NatureCounts and land cover data.

```{r}
# Match the CRS
mauricie_boundary <- st_transform(mauricie_boundary, crs = st_crs(landcover)) 

# Plot
ggplot() +
  # Add the raster layer
  geom_raster(data = landcover_mask_df, aes(x = x, y = y, fill = landcover)) +
  
  # Add the National Park boundary in red
  geom_sf(data = mauricie_boundary, fill = NA, color = "purple", size = 0.9) +
  
  # Add the multipoints from the site summary, sized by total_count
  geom_sf(data = mauricie_site_summary, aes(size = total_count), color = "grey", shape = 21) +
  
  # Custom color scale for landcover
  scale_fill_manual(
    values = setNames(
      rgb(landcover_rgb$red, landcover_rgb$green, landcover_rgb$blue, maxColorValue = 255),
      landcover_rgb$landcover
    ),
    name = "Landcover"
  ) +
  
  # Adjust the size scale for points
  scale_size_continuous(name = "Total Count", range = c(1, 6)) +  # Adjust the range for better visibility
  
  # Add labels and theme
  labs(
    title = "La Mauricie National Park - Landcover 2015",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_bw() +
  
  # Use coord_sf for compatibility with sf layers
  coord_sf()
```

Extract the land cover class id for each point id and append it to `mauricie_birds_df`.

First, make sure the CRS of both spatial data match and then use `terra::extract()`.

```{r}
# Match the CRS
mauricie_birds_sf <- st_transform(mauricie_birds_sf, crs = st_crs(landcover_mask))

# Extract the landcover class ID for each observation
landcover_class_sf <- terra::extract(landcover_mask, vect(mauricie_birds_sf), bind = TRUE)
```

Convert the sf object holding the extracted land cover class ID's to a dataframe. Use `terra::as.data.frame()` explicitly when calling the function to avoid error.

```{r}
landcover_class_df <- terra::as.data.frame(landcover_class_sf) %>%
  select(point_id, landcover_class = Class) %>%
  mutate(landcover_class = as.character(landcover_class)) %>%  # Convert to character
  left_join(
    landcover_rgb %>% 
      mutate(landcover_class = as.character(landcover_class)), # Convert to character
    by = "landcover_class"
  )
```

OPTIONAL: To save the land cover data as a *.csv* to your disk, use the `write.csv()` function, specify the name of your .*csv* file, and use the row.names = FALSE argument to exclude row numbers from the output. **The extracted landcover values will be required to complete [Chapter 7: Raster Summary Tools](07-RasterSummaryTools.Rmd)**. We have uploaded this file to the [**Git repository data folder**](https://github.com/BirdsCanada/NatureCounts_SpatialData_Tutorial/tree/d83eed714373048c8f9fdb4ea4a17f1fd4bef3f7/Data) and the [**Google Drive data folder**](https://drive.google.com/drive/folders/1gLUC6fROl4kNBvTGselhZif-arPexZbY?usp=sharing) for your convenience.

> To execute this code chunk, remove the \#

```{r}
# write.csv(landcover_class_df, "data/mauricie/landcover_class_df.csv", row.names = FALSE)
```

Combine the color table and land cover class dataframes with the NatureCounts dataframe.

```{r}
mauricie_birds_df <- mauricie_birds_df %>%
  left_join(
    landcover_class_df %>%
      mutate(landcover_class = as.character(landcover_class)), # Convert to character
    by = "record_id" # join by record_id
  ) %>%
  left_join(
    landcover_rgb %>%
      mutate(landcover_class = as.character(landcover_class)), # Convert to character
    by = "landcover_class" # join by land cover class
  )

```

------------------------------------------------------------------------

**Congratulations**! You completed **Chapter 5 - Land Cover Data**. In this chapter, you successfully plotted land cover (raster) data, interpreted a raster color table, and combined NatureCounts data with land cover class values. In [Chapter 6](06-SatelliteImagery.Rmd), you can explore how to download satellite imagery and calculate spectral indices to combine with NatureCounts data.
