---
title: "05-LandcoverData"
author: "Dimitrios Markou"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chapter 5: Land Cover Data

##### Authors: Dimitrios Markou, Danielle Ethier

> In [Chapter 4](04-ElevationData.Rmd), you processed Digital Terrain Models, applied crop and mask procedures, and extracted elevation values to combine with NatureCounts data. In this tutorial, you will extract land cover data over [La Mauricie National Park](https://parks.canada.ca/pn-np/qc/mauricie/nature) and combine these values to NatureCounts data from the [Quebec Breeding Bird Atlas (2010 - 2014)](https://naturecounts.ca/nc/default/datasets.jsp?code=QCATLAS2PC&sec=bmdr).

**Land cover** describes the surface cover on the ground like vegetation, urban, water, and bare soil while **land use** describes the purpose that the land serves like recreation, wildlife habitat, and agriculture. See [Land Cover & Land Use](https://natural-resources.canada.ca/maps-tools-and-publications/satellite-imagery-elevation-data-and-air-photos/tutorial-fundamentals-remote-sensing/educational-resources-applications/land-cover-land-use/land-cover-land-use/9373) from Natural Resources Canada for more information.

**To proceed with this tutorial, download the necessary packages in [5.0 Learning Objectives] and complete section 4.1 Data Setup from [Chapter 4: Elevation Data](04-ElevationData.Rmd)**. **It helps you download the National Park boundary and NatureCounts data which are required for this lesson. Alternatively, you can access these files in the [Git repository data folder](https://github.com/BirdsCanada/NatureCounts_SpatialData_Tutorial/tree/d83eed714373048c8f9fdb4ea4a17f1fd4bef3f7/Data) or the [Google Drive data folder](https://drive.google.com/drive/folders/1gLUC6fROl4kNBvTGselhZif-arPexZbY?usp=sharing). Make sure to grab the *.shp* file as well as ALL accompanying files.**

# 5.0 Learning Objectives

By the end of **Chapter 5 - Land Cover Data**, users will know how to:

-   Load and process land cover data (raster), including cropping and masking
-   Extract unique pixel values from a raster data set
-   Combine NatureCounts data with land cover data for analysis

This R tutorial requires the following packages:

```{r, eval = TRUE, warning = FALSE, message = FALSE}
library(naturecounts)
library(sf)
library(terra)
library(tidyverse)
library(lubridate)
library(readr)
library(leaflet)
library(leaflet.extras)
```

This tutorial uses the following spatial data.

1.  [Places administered by Parks Canada](https://open.canada.ca/data/en/dataset/e1f0c975-f40c-4313-9be2-beb951e35f4e/resource/0744186a-449b-4f1f-9f1e-952a94c6d5ca) - Boundary shapefiles\*

2.  [Quebec Breeding Bird Atlas (2010 - 2014)](https://naturecounts.ca/nc/default/datasets.jsp?code=QCATLAS2PC&sec=bmdr) - NatureCounts bird observations\*

    \*These data are available for download via Section 4.1: Data Setup in [Chapter 4](04-ElevationData.Rmd), the [Git repository data folder](https://github.com/BirdsCanada/NatureCounts_SpatialData_Tutorial/tree/d83eed714373048c8f9fdb4ea4a17f1fd4bef3f7/Data) or the [Google Drive data folder](https://drive.google.com/drive/folders/1gLUC6fROl4kNBvTGselhZif-arPexZbY?usp=sharing)

3.  [2015 Land Cover of Canada](https://open.canada.ca/data/en/dataset/4e615eae-b90c-420b-adee-2ca35896caf6) - Land Cover map (30 m resolution)

# 5.1 Land Cover Data

Read in the National Park boundary you saved or downloaded from [Chapter 4: Elevation Data](04-ElevationData.Rmd).

```{r}
mauricie_boundary <- st_read("Data/mauricie/boundary/mauricie_boundary.shp")
```

To read in the land cover dataset, navigate to [2015 Land Cover of Canada](https://open.canada.ca/data/en/dataset/4e615eae-b90c-420b-adee-2ca35896caf6), scroll down to **Data and Resources** and select the TIFF file to download. Extract the data download and save the file to your directory before applying `terra::rast()`.

```{r}
# if the data are in your working directory
 landcover <- rast("landcover-2015-classification.tif")

# else, specify the location of your data
# landcover <- rast(path/to/your/landcover-2015-classification.tif)

print(landcover)
```

Transform the CRS of the National Park boundary to match that of the land cover dataset.

```{r}
mauricie_boundary <- st_transform(mauricie_boundary, crs = st_crs(landcover))
```

# 5.2 Cropping and masking

**Cropping** reduces the extent of a raster to the extent of another raster or vector.

To crop a raster we can apply the `crop()` function from the `terra` package which uses the SpatVector format. Here, we crop the extent of the raster while converting the National Park Boundary to a SpatVector using `vect()`.
Crop the national dataset to reduce the extent of the raster to the extent of the National Park.

```{r}
landcover_crop <- crop(landcover, vect(mauricie_boundary))
print(landcover_crop)
```


**Masking** assigns NA values to cells of a raster not covered by a vector.

To mask a raster to a vector extent we can apply the `mask()` function from the `terra` package which also uses the SpatVector format. 
Mask the national dataset and apply NA values to all cells outside the extent of the National Park.

```{r}
landcover_mask <- mask(landcover_crop, vect(mauricie_boundary))
print(landcover_mask)
```

```{r}
# Set up the plot area for 1 row and 2 columns
par(mfrow = c(1, 2))

# Plot the cropped land cover raster
plot(landcover_crop, main = "crop")

# Add National Park boundary
plot(st_geometry(mauricie_boundary), col = NA, border ="purple", lwd = 2, add = TRUE)

# Plot the masked land cover raster 
plot(landcover_mask, main = "mask")

# Add National Park boundary
plot(st_geometry(mauricie_boundary), col = NA, border ="purple", lwd = 2, add = TRUE)

# Reset plotting layout to default
par(mfrow = c(1, 1))
```

OPTIONAL: To write the masked raster to your disk, you can use the writeRaster() function from terra. **This raster will be required to complete [Chapter 7: Raster Summary Tools](07-RasterSummaryTools.Rmd)**. You must complete the above sections and save this raster to proceed with Chapter 7 unless you're using your own data.

To execute this code chunk, remove the \#

```{r}

# getwd() # to get the path to your working directoy

# output_path <- "path/to/workingdirectory/" 

# writeRaster(landcover_mask, output_path, overwrite = TRUE)
```

Examine the contents of the masked land cover raster. The `freq()` function describes the pixel count per land cover class represented in the region.

```{r}
freq(landcover_mask)
```

Examine the categories of the masked land cover raster. The `cats()` function displays the color (RGB) table that associates each class with a distinct color. Let's save this to our environment as a dataframe called **landcover_rgb** using the `as.data.frame()` function.

```{r}
landcover_rgb <- as.data.frame(coltab(landcover_mask))
landcover_rgb
```

Filter the RGB dataframe to only include those classes represented in the masked land cover raster. To assign the appropriate labels to each land cover class, refer to the [Class Index](https://open.canada.ca/data/en/dataset/ee1580ab-a23d-4f86-a09b-79763677eb47/resource/b8411562-49b7-4cf6-ac61-dbb893b182cc) and use the `mutate()` function after filtering for the relevant classes and tidying-up the attribute names.

```{r}
landcover_rgb <- landcover_rgb %>%
  rename_with(tolower) %>% # lowercase is more R friendly!
  rename(landcover_class = value) %>%
  filter(landcover_class %in% c(1, 5, 6, 8, 10, 14, 16, 17, 18)) %>% # filter based on the classes represented in the "landcover_crop"
  mutate(landcover = c("needleleaf", "broadleaf", "mixed forest", "shrubland", "grassland", "wetland", "barren", "urban", "water")) # assign land cover class labels
```

Let's combine the RGB values with the raster data.

First, convert the cropped landcover raster to a dataframe.

```{r}
landcover_mask_df <- as.data.frame(landcover_mask, xy = TRUE, cells = TRUE)
```

Next, merge the two dataframes by performing a `left_join()`. Ensure that the grouping variable for both dataframe matches.

```{r}
landcover_mask_df <- landcover_mask_df %>%
  rename(landcover_class = Class) %>%
  left_join(landcover_rgb %>% mutate(landcover_class = as.factor(landcover_class)), by = "landcover_class")
```

Map the cropped land cover raster using `ggplot()`.

```{r}
ggplot() +
  geom_raster(data = landcover_mask_df, aes(x = x, y = y, fill = landcover)) +
  scale_fill_manual(
    values = setNames(
      rgb(landcover_rgb$red, landcover_rgb$green, landcover_rgb$blue, maxColorValue = 255),
      landcover_rgb$landcover
    ),
    name = "Landcover"
  ) +
  labs(
    title = "La Mauricie National Park - Landcover 2015",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_bw() +
  coord_fixed() +
  geom_sf(data = mauricie_boundary, fill = NA, color = "purple", size = 1.5)
```

Read in the NatureCounts data you saved or downloaded from [Chapter 4: Elevation Data](04-ElevationData.Rmd).

```{r, warning = FALSE, message = FALSE}
mauricie_birds_df <- read_csv("data/mauricie/mauricie_birds_df.csv") 
```

To convert the NatureCounts data to a spatial object and transform its CRS to match the National Park boundary we can use the `st_as_sf()` and `st_transform()` functions, respectively.

```{r}
mauricie_birds_sf <- sf::st_as_sf(mauricie_birds_df, 
                        coords = c("longitude", "latitude"), crs = 4326)              # convert to sf object

mauricie_boundary <- st_transform(mauricie_boundary, crs = st_crs(mauricie_birds_sf)) # match the CRS
```

Summarize the NatureCounts data by site for mapping.

```{r}
# Group by SiteCode and summarize total_count
mauricie_site_summary <- mauricie_birds_sf %>%
  group_by(SiteCode) %>%
  summarize(total_count = sum(ObservationCount, na.rm = TRUE)) 
```

Now we can map the NatureCounts and land cover data.

```{r}
# Match the CRS
mauricie_boundary <- st_transform(mauricie_boundary, crs = st_crs(landcover)) 

# Plot
ggplot() +
  # Add the raster layer
  geom_raster(data = landcover_mask_df, aes(x = x, y = y, fill = landcover)) +
  
  # Add the National Park boundary in red
  geom_sf(data = mauricie_boundary, fill = NA, color = "purple", size = 0.9) +
  
  # Add the multipoints from the site summary, sized by total_count
  geom_sf(data = mauricie_site_summary, aes(size = total_count), color = "yellow", shape = 21) +
  
  # Custom color scale for landcover
  scale_fill_manual(
    values = setNames(
      rgb(landcover_rgb$red, landcover_rgb$green, landcover_rgb$blue, maxColorValue = 255),
      landcover_rgb$landcover
    ),
    name = "Landcover"
  ) +
  
  # Adjust the size scale for points
  scale_size_continuous(name = "Total Count", range = c(1, 6)) +  # Adjust the range for better visibility
  
  # Add labels and theme
  labs(
    title = "La Mauricie National Park - Landcover 2015",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_bw() +
  
  # Use coord_sf for compatibility with sf layers
  coord_sf()
```

Extract the land cover class ID for each observation site and append it to `mauricie_birds`.

First, make sure the CRS of both spatial data match and then use `terra::extract()`.

```{r}
# Match the CRS
mauricie_birds_sf <- st_transform(mauricie_birds_sf, crs = st_crs(landcover_mask))

# Extract the landcover class ID for each observation
landcover_class_sf <- terra::extract(landcover_mask, vect(mauricie_birds_sf), bind = TRUE)
```

Convert the sf object holding the extracted land cover class ID's to a dataframe. Use `terra::as.data.frame()` explicitly when calling the function to avoid error.

##Error generated here. I have not finished reviewing this Chapters
```{r}
landcover_class_df <- terra::as.data.frame(landcover_class_sf) %>% 
 select(record_id, landcover_class = Class) %>%
 left_join(landcover_rgb[, c("landcover_class", "landcover")], by = "landcover_class")
```

OPTIONAL: To save the land cover data as a *.csv* to your disk, use the `write.csv()` function, specify the name of your .csv file, and use the row.names = FALSE argument to exclude row numbers from the output. **The extracted landcover values will be required to complete [Chapter 7: Raster Summary Tools](07-RasterSummaryTools.Rmd) and subsequent chapters**. We have uploaded this file to the [**Git repository data folder**](https://github.com/BirdsCanada/NatureCounts_SpatialData_Tutorial/tree/d83eed714373048c8f9fdb4ea4a17f1fd4bef3f7/Data) **and the [Google Drive data folder](https://drive.google.com/drive/folders/1gLUC6fROl4kNBvTGselhZif-arPexZbY?usp=sharing)** for your convenience.

> To execute this code chunk, remove the \#

```{r}
# write.csv(landcover_class_df, "data/mauricie/landcover_class_df.csv", row.names = FALSE)
```

Combine the land cover dataframe with the NatureCounts dataframe.

```{r}
mauricie_birds_df <- mauricie_birds_df %>%
  left_join(landcover_class_df, by = "record_id") %>% # Join by record_id
  left_join(landcover_rgb, by = "landcover_class") # Join by land cover class
```

**Congratulations**! You completed **Chapter 5 - Land Cover Data**. In this chapter, you successfully plotted land cover (raster) data, extracted unique pixel values, and combined them with NatureCounts data. In [Chapter 6](06-SatelliteImagery.Rmd), you can explore how to download satellite imagery and calculate spectral indices to combine with NatureCounts data.
