---
title: "07-RasterSummaryTools"
author: "Dimitrios Markou"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chapter 7: Raster Summary Tools

##### Author: Dimitrios Markou

> In Chapter 6, you downloaded satellite imagery from the Copernicus SENTINEL-2 mission and calculated spectral indices (NDWI, NDVI) over [La Mauricie National Park](https://parks.canada.ca/pn-np/qc/mauricie/nature) to combine with NatureCounts data from the [Quebec Breeding Bird Atlas (2010 - 2014)](https://naturecounts.ca/nc/default/datasets.jsp?code=QCATLAS2PC&sec=bmdr). In this tutorial, you will combine and reclassify raster data, calculate summary statistics, and create data summaries and visualizations.

**The National Park boundary and NatureCounts data are available for download via the [Git repository](https://github.com/BirdsCanada/NatureCounts_SpatialData_Tutorial/tree/d83eed714373048c8f9fdb4ea4a17f1fd4bef3f7/Data) or [Google Drive](https://drive.google.com/drive/folders/1gLUC6fROl4kNBvTGselhZif-arPexZbY?usp=sharing) data folders. Quick access to the raster data outputs from [Chapter 4: Elevation Data](04-ElevationData.Rmd) and [Chapter 5: Land Cover Data](05-LandcoverData.Rmd) which are required for this lesson are also available through the Google Drive folder. Once there, download and extract the** *dtm_mask (.zip)* **and** *landcover_mask* *(.zip)* **files to your directory. If you wish to gain experience in how to download, process, and save either of these environmental layers yourself, return to the earlier chapters of this tutorial series.**

# 7.0 Learning Objectives {#7.0LearningObjectives}

By the end of **Chapter 7 - Raster Summary Statistics**, users will know how to:

-   Transform and combine raster data objects: [7.1 Raster Data Transformations](#7.1RasterDataTransformations)

-   Reclassify raster data: [7.2 Raster Reclassification](#7.2RasterReclassification)

-   Perform raster algebra using local, global, and zonal functions: [7.3 Raster Algebra](#7.3RasterAlgebra)

-   Create and visualize NatureCounts data summaries using environmental covariates: [7.4 Summary Tools: Species Abundance](#7.4SummaryTools)

This R tutorial requires the following packages:

```{r, eval = TRUE, warning = FALSE, message = FALSE}
library(sf)
library(terra)
library(tidyverse)
library(dplyr)
```

Use `st_read()` from the `sf` function to bring in the **mauricie_boundary** found in the `data` folder of the project repo, if necessary.

```{r}
mauricie_boundary <- sf::st_read("data/mauricie/boundary/mauricie_boundary.shp")
```

# 7.1 Raster Data Transformations {#7.1RasterDataTransformations}

Raster data is stored as a SpatRaster object and can have one or multiple layers. To open a SpatRaster, we can use the `rast()` function from the `terra` package. In [Chapter 4: Elevation Data](04-ElevationData.Rmd), you created a mask of La Mauricie National Park using single-layer DTM elevation data. Read this file from your directory and call it `elevation_rast`:

```{r}
elevation_rast <- rast("path/to/your/dtm_mask.tif")
elevation_rast
```

The function `names()` can be used to rename the layer(s) of a SpatRaster.

```{r}
names(elevation_rast) <- "elevation"
elevation_rast
```

In [Chapter 5: Land Cover Data](05-LandcoverData.Rmd), you created a landcover mask of La Mauricie National Park using the [2015 Land Cover of Canada](https://open.canada.ca/data/en/dataset/4e615eae-b90c-420b-adee-2ca35896caf6) dataset. Read this file from your directory and call it `landcover_rast`:

```{r}
landcover_rast <- rast("path/to/your/landcover_mask.tif")
landcover_rast
```

Rename the land cover class layer.

```{r}
names(landcover_rast) <- "landcover_class"
landcover_rast
```

You'll notice that the resolution, extent, and CRS of the elevation and land cover raster layers do not match which prevents us from computing these layers together. Use the `project()` function from `terra` to transform the coordinate reference system of `elevation_rast` to match that of `landcover_rast` (Canada Atlas Lambert, EPSG: 3979). By default, the "bilinear" method will be used for numeric layers like elevation. Run [help("project")]{.underline} in your **Console** for more details. **This could take about 15 min - stretch break!**.

```{r, eval = true, message = FALSE}
elevation_rast_proj <- terra::project(elevation_rast, "EPSG:3979", method = "bilinear")
```

To match the extents of both layers, crop the `landcover_rast` to the extent of the `elevation_rast_proj`.

```{r}
landcover_rast <- crop(landcover_rast, ext(elevation_rast_proj))
```

Resample `landcover_rast` to match the resolution of `elevation_rast_proj` using the **"near"** method. This is the default method used for categorical layers like land cover. Run [help("resample")]{.underline} in your **Console** for more details.

```{r}
landcover_rast_resample <- resample(landcover_rast, elevation_rast_proj, method = "near")
```

Let's verify that our transformations were successful. The `all()` function will help evaluate a set of conditions and produce a TRUE or FALSE result across all our checks for spatial alignment.

```{r}
all(
  crs(landcover_rast_resample) == crs(elevation_rast_proj), # check crs
  ext(landcover_rast_resample) == ext(elevation_rast_proj), # check extent
  res(landcover_rast_resample) == res(elevation_rast_proj) # check resolution
)
```

Let's test plot the layers side by side.

```{r, eval = TRUE, warning = FALSE}
# Match the CRS
mauricie_boundary <- st_transform(mauricie_boundary, crs = st_crs(landcover_rast))

# Set up a 1x2 plotting layout
par(mfrow = c(1, 2))

# Plot elevation
plot(elevation_rast_proj, axes = TRUE, plg = list(title = ""), main = "Elevation (m)")

# Overlay the Mauricie boundary on the first plot
plot(mauricie_boundary, add = TRUE, border = "red", col = NA, lwd = 2)

# Plot landcover
plot(landcover_rast_resample, axes = TRUE, plg = list(title = ""), main = "Land Cover Class")

# Overlay the Mauricie boundary on the second plot
plot(mauricie_boundary, add = TRUE, border = "red", col = NA, lwd = 2)
```

Multiple single-layer SpatRaster with the same CRS, extent, and resolution, can be combined into a single multi-layer SpatRaster with the `c()` function.

```{r}
mauricie_rast <- c(elevation_rast_proj, landcover_rast_resample)
mauricie_rast
```

You can subset a multi-layer SpatRaster with the \$ or [[]] operators.

```{r}
mauricie_rast$elevation

mauricie_rast[["landcover_class"]]
```

# 7.2 Raster Reclassification {# 7.2RasterReclassification}

When communicating our data, it might be useful to apply **local** operations such as the reclassification of a range of raster values into a smaller set of classes. In this process, a range of values are binned to a new (categorical) value. Reclassifying rasters is easily achieved using `terra` and the `classify()` function.

First, provide a reclassification table in the form of a matrix which specifies the range of raster values (columns 1 and 2) and the class ID (column 3) like so:

```{r}
dtm_rcl_matrix <- matrix(c(0, 200, 1,
                           200, 400, 2,
                           400, 1000, 3), byrow = TRUE, ncol = 3)
```

Then, apply the `classify()` function using the reclassification matrix:

```{r}
elevation_rast_reclass <- classify(elevation_rast_proj, dtm_rcl_matrix)
```

Finally, give meaningful labels to each of the classes:

```{r, eval = TRUE}
elevation_levels <- data.frame(ID = c(1L, 2L, 3L),
                            category = c("Low",
                                         "Mid",
                                         "High"))
levels(elevation_rast_reclass) <- elevation_levels
```

Visualize the regional DTM and National Park boundary using the reclassification scheme:

```{r, eval = TRUE, warning = FALSE}
# Set up a 1x2 plotting layout
par(mfrow = c(1, 2))  # Set up a 1-row, 2-column plotting layout

# Plot the original DTM mosaic
plot(elevation_rast_proj, main = "Regional DTM")

# Overlay the Mauricie boundary on the first plot
plot(mauricie_boundary, add = TRUE, border = "red", col = NA, lwd = 2)

# Plot the reclassified DTM mosaic
plot(elevation_rast_reclass, main = "Reclassified DTM")

# Overlay the Mauricie boundary on the second plot
plot(mauricie_boundary, add = TRUE, border = "red", col = NA, lwd = 2)

# Reset the plotting layout to the default
par(mfrow = c(1, 1))
```

# 7.3 Raster Algebra (Local, Global, and Zonal Operations) {#7.3RasterAlgebra}

Map algebra comprises a slew of useful operations that can be used to modify or summarize raster cell values with reference to nearby cells, zones, or statistical functions.

**Local** operations involve all cell by cell operations across single or multi-layer rasters. In Section [7.2 Raster Reclassification], we performed a local operation by reclassifying elevation raster values into distinct groups. Custom functions used to calculate spectral indices (see [Chapter 6: SatelliteImagery](06-SatelliteImagery.Rmd)) are another example of local (pixel by pixel) operations.

**Global** operations can be used to derive descriptive raster statistics and summaries across entire SpatRaster. For example, you can visualize your raster data values using a histogram (`hist()`).

```{r, warning = FALSE}
hist(mauricie_rast$elevation, axes = TRUE, 
            main = "Elevation values within La Mauricie National Park",
            xlab = "Elevation (m)",
            col = "Green")
```

You can produce global summary statistics for individual SpatRaster layers using the `global()` function. It helps calculate `min()`, `max()`, `mean()`, and `sd()` by specifying the `fun` argument.

```{r}
global_elevation_stats <- global(elevation_rast, fun = c("min", "max", "mean", "sd"), na.rm = TRUE)
print(elevation_stats)
```

You can also use custom functions to calculate global raster summary statistics. For example, `my_summary_stats` calculates the min, max, mean, and sd of an input numeric object `x` and returns the summary statistics in a vector (**give this a few minutes**):

```{r}
my_summary_stats <- function(x) {
  out <- c(min = min(x, na.rm = TRUE),
           max = max(x, na.rm = TRUE),
           mean = mean(x, na.rm = TRUE),
           sd = sd(x, na.rm = TRUE))
  return(out)
}
global(elevation_rast, fun = my_summary_stats) # apply the function
```

**Zonal** operations apply an aggregation function to multiple raster cells. Typically, a second (categorical) raster defines the zonal filters (or 'zones'). To calculate the mean elevation per individual land cover class, we can apply the `zonal()` function.

```{r}
zonal_elevation_stats <- zonal(elevation_rast_proj, landcover_rast_resample, fun = "mean", na.rm = TRUE)
zonal_elevation_stats
```

# 7.4 Summary Tools: Species Abundance {#7.4SummaryTools}

In Chapter 4, Chapter 5, and Chapter 6 you extracted elevation, land cover, and NDVI values, respectively over bird observation sites across La Mauricie National Park. These data were uploaded to the [Git repository data folder](https://github.com/BirdsCanada/NatureCounts_SpatialData_Tutorial/tree/d83eed714373048c8f9fdb4ea4a17f1fd4bef3f7/Data) and the [Google Drive data folder](https://drive.google.com/drive/folders/1gLUC6fROl4kNBvTGselhZif-arPexZbY?usp=sharing) for your convenience.

Let's download all the environmental covariates into R and join them to a common dataframe.

```{r, message = FALSE}
# List the dataframes
env_covariates <- list.files(path = "data/mauricie/env_covariates", 
                             pattern = "\\.csv$", 
                             full.names = TRUE)

# Read each CSV into a list of dataframes
env_covariates_list <- lapply(env_covariates, read_csv)

# Combine NatureCounts and environmental covariates 
env_covariates_df <- Reduce(function(x, y) left_join(x, y, by = "record_id"), env_covariates_list)
```

Read in the NatureCounts data you saved or downloaded from [Chapter 4: Elevation Data](04-ElevationData.Rmd).

```{r, warning = FALSE, message = FALSE}
mauricie_birds_df <- read_csv("data/mauricie/mauricie_birds_df.csv") 
```

Combine the environmental covariates with the NatureCounts data.

```{r}
mauricie_data <- mauricie_birds_df %>%
  merge(env_covariates_df, by = "record_id")
```

We can summarize the combined NatureCounts and environmental data to explore possible trends in species abundance.

#### EXAMPLE 1: Landscape Association Plots

Calculate the species abundance (number of individuals) for each land cover class relative to the entire National Park population.

```{r}
landcover_summary <- mauricie_data %>%
  group_by(landcover) %>%
  summarize(
    total_individuals = sum(ObservationCount, na.rm = TRUE),
    total_species = n_distinct(scientific_name)
  ) %>%
  mutate(mean_abundance = ((total_individuals / sum(total_individuals))))
```

Plot the relative abundance per land cover class.

```{r}
ggplot(landcover_summary, aes(x = landcover, y = mean_abundance, fill = total_individuals)) +
  geom_bar(stat = "identity", color = "black") +
  geom_text(
    aes(label = paste0("n = ", total_individuals)),
    vjust = -0.5, # Position the text above the bars
    size = 3.5    # Adjust text size
  ) +
  scale_fill_gradient(
    low = "lightblue", high = "darkblue", # Customize colors for the gradient
    name = "Abundance"                      # Legend title
  ) +
  theme_minimal() +
  labs(
    title = "Relative Species Abundance by Landcover Class",
    x = "Landcover Class",
    y = "Relative Abundance"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

#### EXAMPLE 2: Species Rank Plot

Group the NatureCounts data by species and rank them in order of abundance.

```{r}
# Filter for needleleaf and broadleaf landcover classes, then group and rank
species_rank_landcover <- mauricie_data %>%
  filter(landcover %in% c("needleleaf", "broadleaf")) %>%
  group_by(landcover, scientific_name) %>%
  summarize(total_abundance = sum(ObservationCount, na.rm = TRUE), .groups = "drop") %>%
  arrange(landcover, desc(total_abundance)) %>%
  group_by(landcover) %>%
  mutate(rank = row_number())
```

Plot the abundance for each species rank across both land cover classes using the sample.

```{r}
ggplot(species_rank_landcover, aes(x = as.numeric(rank), y = total_abundance, color = landcover)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title = "Abundance of High Rank Species across Broadleaf and Needleleaf Forests",
    x = "Species Rank",
    y = "Abundance",
    color = "Landcover"
  ) +
  facet_wrap(~ landcover, scales = "free_x") +  # Separate plots for each landcover
  scale_color_manual(values = c("needleleaf" = "darkgreen", "broadleaf" = "brown")) +
  scale_x_continuous(breaks = seq(1, max(species_rank_landcover$rank), by = 4)  # Set breaks at each rank
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for better readability
    panel.grid.minor = element_blank()
  )

```

List the top 5 species from each land cover type.

```{r}
# List top 5 species by total abundance for each land cover type
top_five <- species_rank_landcover %>%
  group_by(landcover) %>%
  slice_head(n = 5) %>%  # Select the top 5 species for each landcover
  arrange(landcover, desc(total_abundance))  # Arrange by abundance within each landcover class

# View the result
top_five
```

#### EXAMPLE 3: Elevation Plot

```{r}
# Create elevation classes with labels 
mauricie_data <- mauricie_data %>%
  mutate(
    elevation_class = case_when(
      elevation < 200 ~ "Low",
      elevation >= 200 & elevation < 400 ~ "Mid",
      elevation >= 400 ~ "High"
    ),
    elevation_class = factor(elevation_class, levels = c("Low", "Mid", "High"))  # Set the factor levels
  )

# Calculate relative abundance for each elevation class
elevation_summary <- mauricie_data %>%
  group_by(elevation_class) %>%
  summarize(
    total_individuals = sum(ObservationCount, na.rm = TRUE),
    total_species = n_distinct(scientific_name)
  ) %>%
  mutate(
    relative_abundance = total_individuals / sum(total_individuals) 
  )

# Plot relative species abundance per elevation class
ggplot(elevation_summary, aes(x = elevation_class, y = relative_abundance, fill = total_individuals)) +
  geom_bar(stat = "identity", color = "black") +
  geom_text(
    aes(label = paste0("n = ", total_individuals)),
    vjust = -0.5, # Position the text above the bars
    size = 3.5    # Adjust text size
  ) +
  scale_fill_gradient(
    low = "pink", high = "darkred", # Customize colors for the gradient
    name = "Abundance"              # Legend title
  ) +
  theme_minimal() +
  labs(
    title = "Relative Species Abundance by Elevation Class",
    x = "Elevation",
    y = "Relative Abundance"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

#### EXAMPLE 4: NDVI Plots

Plot the relative species abundance for different NDVI ranges.

```{r}
# Create NDVI classes with numeric labels
mauricie_data <- mauricie_data %>%
  mutate(
    ndvi_range = case_when(
      ndvi > 0 & ndvi <= 0.2 ~ "0 to 0.2",
      ndvi > 0.2 & ndvi <= 0.4 ~ "0.2 to 0.4",
      ndvi > 0.4 & ndvi <= 0.6 ~ "0.4 to 0.6",
      ndvi > 0.6 & ndvi <= 0.8 ~ "0.6 to 0.8",
    )
  )

# Calculate species abundance for each NDVI class
ndvi_summary <- mauricie_data %>%
  group_by(ndvi_range) %>%
  summarize(
    total_individuals = sum(ObservationCount, na.rm = TRUE),
    total_species = n_distinct(scientific_name)
  ) %>%
  mutate(
    relative_abundance = total_individuals / sum(total_individuals)
  )

# Plot relative species abundance per NDVI class
ggplot(ndvi_summary, aes(x = ndvi_range, y = relative_abundance, fill = total_individuals)) +
  geom_bar(stat = "identity", color = "black") +
  geom_text(
    aes(label = paste0("n = ", total_individuals)),
    vjust = -0.5, # Position the text above the bars
    size = 3.5    # Adjust text size
  ) +
  scale_fill_gradient(
    low = "lightgreen", high = "darkgreen", # Customize colors for the gradient
    name = "Abundance"                      # Legend title
  ) +
  theme_minimal() +
  labs(
    title = "Relative Species Abundance According to NDVI Distribution",
    x = "NDVI",
    y = "Relative Abundance"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Plot species richness and minimum NDVI grouped by SiteCode.

```{r}
# Calculate species richness by SiteCode
species_richness <- mauricie_data %>%
  group_by(SiteCode) %>%
  summarise(species_richness = n_distinct(scientific_name))

# Calculate minimum NDVI by SiteCode
min_ndvi <- mauricie_data %>%
  group_by(SiteCode) %>%
  summarise(min_ndvi = min(ndvi, na.rm = TRUE))

# Merge species richness and mean NDVI by SiteCode
species_ndvi_data <- left_join(species_richness, min_ndvi, by = "SiteCode")

# Plot species richness and minimum NDVI
ggplot(species_ndvi_data, aes(x = min_ndvi, y = species_richness)) +
  geom_point() +
  labs(title = "Species Richness and Minimum NDVI at Each Observation Site",
       x = "Min NDVI",
       y = "Bird Species Richness") +
  theme_minimal()
```

------------------------------------------------------------------------

Congratulations! You completed Chapter 7 - Raster Summary Tools. In this chapter, you successfully transformed and combined raster data, performed raster algebra operations, and created summary plot for NatureCounts data using environmental covariates.
