---
title: "07-SummaryTools"
author: "Dimitrios Markou"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chapter 7: Summary Tools

##### Authors: Dimitrios Markou, Danielle Ethier

> In Chapter 6, you downloaded satellite imagery from the Copernicus SENTINEL-2 mission and calculated spectral indices (NDWI, NDVI) over [La Mauricie National Park](https://parks.canada.ca/pn-np/qc/mauricie/nature) to combine with NatureCounts data from the [Quebec Breeding Bird Atlas (2010 - 2014)](https://naturecounts.ca/nc/default/datasets.jsp?code=QCATLAS2PC&sec=bmdr). In this chapter, you will create data summaries and visualizations using NatureCounts data and environmental covariates.

**This chapter uses the data products prepared in Chapters 4-6 of the Spatial_Data_Tutorial, and the National Park boundary and NatureCounts data downloaded in section 4.1 Data Setup from [Chapter 4: Elevation Data](04-ElevationData.Rmd). For quick access, all data are available for download via the [Google Drive](https://drive.google.com/drive/folders/1gLUC6fROl4kNBvTGselhZif-arPexZbY?usp=sharing) data folder. If you wish to gain experience in how to download, process, and save the environmental layers yourself, return to the earlier chapters of this tutorial series and explore the Additional Resources articles.**

# 7.0 Learning Objectives {#7.0LearningObjectives}

By the end of **Chapter 7 - Summary Statistics**, users will know how to:

-   Create and visualize NatureCounts and environmental data summaries using four key examples: Landscape Association Plot, Species Rank Plot, Elevation Plot, and NDVI Plot

Load the required packages:

```{r, eval = TRUE, warning = FALSE, message = FALSE}
library(tidyverse)
library(sf)
library(MASS)
library(corrr)
library(ggcorrplot)
library(FactoMineR)
```

# 7.1 Data Setup {#7.4SummaryTools}

In [Chapter 4](04-ElevationData.Rmd), [Chapter 5](05-LandcoverData.Rmd), and [Chapter 6](06-SatelliteImagery.Rmd) you extracted elevation, land cover, and NDVI values, respectively over bird observation sites across La Mauricie National Park. These data were uploaded to the [Google Drive data folder](https://drive.google.com/drive/folders/1gLUC6fROl4kNBvTGselhZif-arPexZbY?usp=sharing) for your convenience.

Let's download all the environmental covariates into R and join them to a common dataframe.

```{r, message = FALSE}
# List the dataframes
env_covariates <- list.files(path = "data/mauricie/env_covariates", 
                             pattern = "\\.csv$", 
                             full.names = TRUE)

# Read each CSV into a list of dataframes
env_covariates_list <- lapply(env_covariates, read_csv)

# Combine NatureCounts and environmental covariates 
env_covariates_df <- Reduce(function(x, y) left_join(x, y, by = "record_id"), env_covariates_list)
```

Read in the NatureCounts data you saved or downloaded from [Chapter 4: Elevation Data](04-ElevationData.Rmd).

```{r, warning = FALSE, message = FALSE}
mauricie_birds_df <- read_csv("data/mauricie/mauricie_birds_df.csv") 
```

Create an `sf` object from the NatureCounts data that represents the unique point count locations.

```{r}
mauricie_birds_sf <- mauricie_birds_df %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) 
```

Trim the `sf` object down by selecting key attribute columns for better readability. Explicitly call the `dplyr` function to avoid namespace clashes. 
```{r}
mauricie_birds_sf <- mauricie_birds_sf %>% 
  dplyr::select(record_id, species_id, SiteCode, Locality, SamplingEventIdentifier, RouteIdentifier, 
         survey_year, survey_month, survey_day, english_name, ObservationCount, geometry)
```

Combine the environmental covariates with the NatureCounts data.

```{r}
mauricie_data <- mauricie_birds_sf %>%
  merge(env_covariates_df, by = "record_id")
```

Assign a point identifier to each location based on its unique geometry.

```{r}
mauricie_data_summary <- mauricie_data %>%
  group_by(SiteCode, geometry) %>%
  mutate(point_id = cur_group_id()) %>%
  ungroup() %>%
  distinct()
```

Convert the summary `sf` object back to a regular dataframe using the `st_drop_geometry()` function.

```{r}
mauricie_data_summary <- st_drop_geometry(mauricie_data_summary)
```

Calculate the species richness and abundance at each point.

```{r}
biodiversity_count <- mauricie_data_summary  %>%
  group_by(point_id) %>%
  summarise(n_species = n_distinct(english_name),
            n_individuals = sum(ObservationCount, na.rm = TRUE), .groups = "drop") # Count unique species
```

Drop all columns except the `point_id` and environmental variables and group all rows together by location.

```{r}
enviro_data_df <- mauricie_data_summary %>%
  dplyr::select(-record_id, -species_id, -english_name, -ObservationCount, -SiteCode, 
         -Locality, -SamplingEventIdentifier, -RouteIdentifier, -survey_year, -survey_month, -survey_day) %>%
  group_by(point_id) %>%
  summarise(across(everything(), ~ first(na.omit(.), default = NA)), .groups = "drop") %>%
  ungroup()
```

Join the `biodiversity_count` dataframe. Cleanup variable names.

```{r}
enviro_data_df <- enviro_data_df %>%
  left_join(biodiversity_count, by = "point_id") %>%
  rename(pland_mixed_forest = `pland_mixed forest`) %>%  # Rename column
  mutate(across(-point_id, as.numeric))  # Ensure all other columns are numeric
```


# 7.2 Diagnostic Plots

Retrieve basic summary info for the environmental data.

```{r}
summary(enviro_data_df)
```

Create a scatterplot matrix to select meaningful variables for modeling. 

```{r}
pairs(~ n_species + n_individuals + elevation + np + ed + pland_mixed_forest + ndvi, data=enviro_data_df, 
      main="Scatterplot Matrix of environmental variables")
```

Inspect the distribution and 'skewness' of the environmental data using a histogram plot.

```{r}
# Select variables and pivot longer for faceting
hist_data <- enviro_data_df %>%
  dplyr::select(n_species, elevation, ed, pland_mixed_forest, ndvi) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# Create faceted histogram plot
ggplot(hist_data, aes(x = value)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  facet_wrap(~ variable, scales = "free") +  # Separate histograms for each variable
  theme_minimal() +
  labs(title = "Multipanel Histogram of Environmental Variables",
       x = "Value",
       y = "Frequency")
```

Create Q-Q plots for each variable to visually assess if the data is normally distributed.  

```{r}
par(mfrow=c(2,3))  # set layout for multiple plots

qqnorm(enviro_data_df$n_species, main = "Q-Q Plot: n_species")
qqline(enviro_data_df$n_species, col = "red")

qqnorm(enviro_data_df$n_individuals, main = "Q-Q Plot: n_individuals")
qqline(enviro_data_df$n_individuals, col = "red")

qqnorm(enviro_data_df$elevation, main = "Q-Q Plot: elevation")
qqline(enviro_data_df$elevation, col = "red")

qqnorm(enviro_data_df$np, main = "Q-Q Plot: np")
qqline(enviro_data_df$np, col = "red")

qqnorm(enviro_data_df$ed, main = "Q-Q Plot: ed")
qqline(enviro_data_df$ed, col = "red")

qqnorm(enviro_data_df$pland_mixed_forest, main = "Q-Q Plot: pland_mixed_forest")
qqline(enviro_data_df$pland_mixed_forest, col = "red")

qqnorm(enviro_data_df$ndvi, main = "Q-Q Plot: ndvi")
qqline(enviro_data_df$ndvi, col = "red")
```

Create a model to describe species richness using elevation, ed, and ndvi.

```{r}
model <- lm(n_species ~  elevation + ed + ndvi , data=enviro_data_df)

summary(model)
anova(model)
```
Create diagnostic plots to evaluate the strength of the model.

```{r}
enviro_data_df$yhat.1 <- fitted(model)  
enviro_data_df$resid.1 <- resid(model) 
enviro_data_df$stdresid.1 <- resid(model)/summary(model)$sigma 

plot(yhat.1 ~ n_species, data=enviro_data_df, xlab="Observed n_species", 
     ylab="Predicted n_species", main="Predicted vs. observed n_species",
     pch=19)
abline(a=0,b=1,col="red")   # plot a reference line 


plot(resid.1 ~ yhat.1, data = enviro_data_df, xlab="Predicted n_species", 
     ylab="Residuals",main="Residual Plot",pch=19)
abline(a=0,b=0,col="red")
```

# 7.2 Species Rank Plots

Species rank plots show the relative abundance (number of individuals) of a species in a community. The number of individuals of each species are sorted in ascending or descending order.

Group the NatureCounts data by species and rank them in order of abundance.

```{r}
species_rank <- mauricie_data %>%
  filter(!is.na(english_name)) %>%  # Remove rows with NA in english_name
  group_by(english_name) %>%  # Group by species only
  summarize(total_abundance = sum(ObservationCount, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_abundance)) %>%  # Sort in descending order of abundance
  slice_max(total_abundance, n = 40) %>%  # Keep only the top 40 species
  mutate(rank = row_number())  # Assign rank to each species
```

Plot the abundance of each species and its rank across the entire park.

```{r}
ggplot(species_rank, aes(x = reorder(english_name, rank), y = total_abundance)) +
  geom_line(group = 1, size = 1, color = "black") +
  geom_point(size = 2, color = "darkgreen") +
  theme_minimal() +
  labs(
    title = "Abundance of High Rank Species across La Mauricie Park",
    x = "Species",
    y = "Total Abundance"
  ) +
  theme(
    axis.text.x = element_text(size = 8, angle = 60, hjust = 1)
  )
```

# 7.4 Data Summary Plots

Create scatter plots with regression lines to explore the relationship of species richness with each environmental variable. 

```{r}
# Elevation vs. n_species
ggplot(enviro_data_df, aes(x = elevation, y = n_species)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Relationship between Elevation and n_species")

# PLAND vs. n_species
ggplot(enviro_data_df, aes(x = pland_mixed_forest, y = n_species)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Relationship between PLAND and n_species")

# Edge Density vs. n_species
ggplot(enviro_data_df, aes(x = ed, y = n_species)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Relationship between Edge Density and n_species")

# NDVI vs. n_species
ggplot(enviro_data_df, aes(x = ndvi, y = n_species)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Relationship between NDVI and n_species")
```

**Conclusion**: There is no clear relationship between species richness and elevation, ED, or NDVI within La Mauricie National Park. **Note**: These results assume that sampling is random with regards to elevation, which is unlikely.

Create a bar plot that represent species richness in binned elevation categories: Low, Mid, and High

```{r}
# Create elevation classes with labels
enviro_data_df <- enviro_data_df %>%
  mutate(
    elevation_class = case_when(
      elevation < 200 ~ "Low",
      elevation >= 200 & elevation < 400 ~ "Mid",
      elevation >= 400 ~ "High"
    ),
    elevation_class = factor(elevation_class, levels = c("Low", "Mid", "High"))  # Set the factor levels
  )

# Summarize species richness per elevation class
elevation_summary <- enviro_data_df %>%
  group_by(elevation_class) %>%
  summarize(spec_richness = sum(n_species, na.rm = TRUE), .groups = "drop")

# Plot species richness per elevation class
ggplot(elevation_summary, aes(x = elevation_class, y = spec_richness)) +
  geom_bar(stat = "identity", color = "black", fill = "steelblue") +
  geom_text(
    aes(label = paste0("n = ", spec_richness)),
    vjust = -0.5, # Position the text above the bars
    size = 3.5    # Adjust text size
  ) +
  theme_minimal() +
  labs(
    title = "Species Richness by Elevation Class",
    x = "Elevation Class",
    y = "Species Richness"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate labels if needed
```

# 7.5 Principal Component Analysis

Null values can bias the result of PCA. In `enviro_data_df`, PLAND null values represent land cover types not represented at a given observation site. Replace the NA's with zeros. 

```{r}
pland_data <- enviro_data_df %>%
  dplyr::select(starts_with("pland_")) %>% # filter for pland variables
  mutate(across(everything(), ~ replace_na(., 0))) %>% # replace NA's withe zero
  scale() # normalize the data 
```

Compute the PCA using the `princomp()` and display the result using `summary()`.

```{r}
pland_data_pca <- princomp(pland_data)
summary(pland_data_pca)
```
Use a scree plot to visualize the importance of each PC.

```{r}
fviz_eig(pland_data_pca, addlabels = TRUE)
```

The first three PC's explain over 70% of the variance. Explore the loadings matrix of the first three principal components. 

```{r}
pland_data_pca$loadings[, 1:3]
```
Create a biplot of the the PLAND variables.

```{r}
fviz_pca_var(pland_data_pca, col.var = "black")
```

```{r}
fviz_cos2(pland_data_pca, choice = "var", axes = 1:2)
```
```{r}
fviz_pca_var(pland_data_pca, col.var = "cos2",
            gradient.cols = c("black", "orange", "green"),
            repel = TRUE)
```

# 7.6  K-Means Clustering

```{r}
# Determine the optimal number of clusters using the elbow method
wss <- sapply(1:10, function(k) kmeans(pland_data, centers = k, nstart = 25)$tot.withinss)

# Plot Elbow Method
plot(1:10, wss, type = "b", pch = 19, frame = FALSE, 
     xlab = "Number of clusters", ylab = "Total Within-Cluster Sum of Squares")

# Choose optimal k
set.seed(123)
kmeans_result <- kmeans(pland_data, centers = 3, nstart = 25)

# Add cluster assignments to the original dataframe
enviro_data_df$cluster <- factor(kmeans_result$cluster)

# **Dimensionality Reduction for Visualization (PCA)**
pca_result <- prcomp(pland_data, center = TRUE, scale. = TRUE)
df_pca <- as.data.frame(pca_result$x[, 1:2])  # Extract first 2 PCs
df_pca$cluster <- enviro_data_df$cluster  # Add cluster info

# Plot clusters in PCA space
ggplot(df_pca, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(size = 3) +
  labs(title = "K-Means Clustering of Species Richness and PLAND metrics",
       x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal()
```
```{r}
ggplot(enviro_data_df, aes(x = cluster, y = pland_mixed_forest, fill = cluster)) +
  geom_boxplot() +
  labs(title = "Forest Cover (%) Across Clusters", y = "PLAND (mixed forest)")
```

------------------------------------------------------------------------

Congratulations! You completed Chapter 7 - Raster Summary Tools. In this chapter, you successfully transformed and combined raster data, performed raster algebra operations, and created summary plot for NatureCounts data using environmental covariates.
