---
title: "07-SummaryTools"
author: "Dimitrios Markou"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chapter 7: Spatial Summary Tools

##### Authors: Dimitrios Markou, Danielle Ethier

> In Chapter 6, you downloaded satellite imagery from the Copernicus SENTINEL-2 mission and calculated spectral indices (NDWI, NDVI) over [La Mauricie National Park](https://parks.canada.ca/pn-np/qc/mauricie/nature) to combine with NatureCounts data from the [Quebec Breeding Bird Atlas (2010 - 2014)](https://naturecounts.ca/nc/default/datasets.jsp?code=QCATLAS2PC&sec=bmdr). In this chapter, you will create data summaries and visualizations using NatureCounts data and environmental covariates.

**This chapter uses the data products prepared in Chapters 4-6, and the National Park boundary and NatureCounts data downloaded in section 4.1 Data Setup from [Chapter 4: Elevation Data](04-ElevationData.Rmd). For quick access, all data are available for download via the [Google Drive](https://drive.google.com/drive/folders/1gLUC6fROl4kNBvTGselhZif-arPexZbY?usp=sharing) data folder. If you wish to gain experience in how to download, process, and save the environmental layers yourself, return to the earlier chapters of this tutorial series and explore the Additional Resources articles.**

# 7.0 Learning Objectives {#7.0LearningObjectives}

By the end of **Chapter 7 - Spatial Summary Tools**, users will know how to:

-   Explore variable relationships and data distribution: [Diagnostics Plots](#7.2DiagnosticPlots)
-   Show the relative abundance (number of individuals) of a species in a community: [Species Rank Plots](#7.3SpeciesRank)
-   Explore the relationship of species richness with each environmental variable: [Regression Plots](#7.4Regression)
-   Visualize species-specific distributions: [Presence/Absence Plots](#7.5PresenceAbsence)

Load the required packages:

```{r, eval = TRUE, warning = FALSE, message = FALSE}
library(tidyverse)
library(sf)
library(MASS)
library(corrr)
library(ggcorrplot)
library(FactoMineR)
library(factoextra)
library(ggplot2)
library(naturecounts)
library(tidyr)
```

This tutorial uses environmental data extracted or calculated from previous chapters of the Spatial Data Tutorial. In [Chapter 5: Land Cover Data](05-LandcoverData), the `landscapemetrics` package was used to calculate some common metrics that help quantify habitat composition and configuration:

-   Percent landcover (PLAND) - percent of landscape of a given class
-   Edge density (ED) - total boundary length of all patches of a given class per unit area
-   Number of patches (NP) - number of unique contiguous units

In [Chapter 6: Satellite Imagery](06-SatelliteImagery.Rmd), the Normalized Difference Vegetation Index (NDVI) was calculated using Sentinel-2 imagery and custom functions .

# 7.1 Data Setup {#7.1DataSetup}

In [Chapter 4](04-ElevationData.Rmd), [Chapter 5](05-LandcoverData.Rmd), and [Chapter 6](06-SatelliteImagery.Rmd) you extracted elevation, land cover, and NDVI values, respectively, at surveys sites across La Mauricie National Park. These data were uploaded to the [Google Drive](https://drive.google.com/drive/folders/1j6PHoUJaFDucWe8V5xXNzdGD00WJu0hq?usp=drive_link) for your convenience. These data should be neatly stored in your `env_covariates` subdirectory.

Run the code chunk below to create your subdirectory, if necessary.

```{r}
if (!dir.exists("data/env_covariates")) { # checks if "env_covariates" subdirectory exists
  print("Subdirectory does not exist. Creating now...")
  dir.create("data/env_covariates", recursive = TRUE) # if not, creates subdirectory
} else {
  print("Subdirectory already exists.")
}
```

Let's download all the environmental covariates and join them to a common dataframe.

```{r, message = FALSE}
# List the dataframes
env_covariates <- list.files(path = "data/env_covariates",
                             pattern = "\\.csv$", 
                             full.names = TRUE)

# Read each CSV into a list of dataframes
env_covariates_list <- lapply(env_covariates, read_csv)

# Combine NatureCounts and environmental covariates 
env_covariates_df <- Reduce(function(x, y) left_join(x, y, by = "point_id"), env_covariates_list)
```

Read in the NatureCounts data you downloaded from [Chapter 4: Elevation Data](04-ElevationData.Rmd) or the [Google Drive](https://drive.google.com/file/d/1ZDo4OUaxqtHuaM1CaGD7BfFGMg1JzOqv/view?usp=drive_link). This file should be in your `data` folder.

```{r, warning = FALSE, message = FALSE}
mauricie_birds_df <- read_csv("data/mauricie_birds_df.csv") 
```

Create an `sf` object from the NatureCounts data that represents the unique point count locations.

```{r}
mauricie_birds_sf <- mauricie_birds_df %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) 
```

Select attribute columns for better readability. Explicitly call the `dplyr` function to avoid namespace clashes.

```{r}
mauricie_birds_sf <- mauricie_birds_sf %>% 
  dplyr::select(SiteCode, survey_year, survey_month, survey_day, english_name, ObservationCount, geometry)
```

Assign a point identifier to each location based on its unique geometry.

```{r}
mauricie_birds_summary <- mauricie_birds_sf %>%
  group_by(SiteCode, geometry) %>%
  mutate(point_id = cur_group_id()) %>%
  ungroup() %>%
  distinct() %>%
  st_drop_geometry(mauricie_birds_summary) # drops geometry and converts to regular dataframe
```

Calculate the species diversity and abundance at each point.

```{r}
biodiversity_count <- mauricie_birds_summary  %>%
  group_by(point_id) %>%
  summarise(n_species = n_distinct(english_name),
            n_individuals = sum(ObservationCount, na.rm = TRUE), .groups = "drop") # Count unique species
```

Join the `env_covariates_df` and `biodiversity_count` dataframes. Cleanup variable names.

```{r}
enviro_data_df <- env_covariates_df %>%
  left_join(biodiversity_count, by = "point_id") %>%
  rename(pland_mixed_forest = `pland_mixed forest`) %>%  # Rename column
  mutate(across(-point_id, as.numeric))  # Ensure all other columns are numeric
```

# 7.2 Diagnostic Plots {#7.2DiagnosticPlots}

Diagnostic plots like summaries, scatterplot matrices, and histograms are essential for evaluating variable relationships, detecting collinearity, and understanding data distributions. They help you understand your data by identifying patterns, outliers, and multicollinearity, ensure accurate model assumptions, and improve data-driven decision-making in future statistical analyses.

Retrieve basic summary info for the environmental data.

```{r}
summary(enviro_data_df)
```

Create a scatterplot matrix to assess the collinearity of the variables.

```{r}
pairs(~ n_species + n_individuals + elevation + np + ed + pland_mixed_forest + ndvi, data=enviro_data_df, 
      main="Scatterplot Matrix of Environmental Variables")
```

Inspect the distribution and 'skewness' of the environmental data using a histogram plot.

```{r}
# Select variables and pivot longer for faceting
hist_data <- enviro_data_df %>%
  dplyr::select(n_species, elevation, ed, pland_mixed_forest, ndvi) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# Create faceted histogram plot
ggplot(hist_data, aes(x = value)) +
  geom_histogram(bins = 10, fill = "skyblue", color = "black") +
  facet_wrap(~ variable, scales = "free") +  # Separate histograms for each variable
  theme_minimal() +
  labs(title = "Multipanel Histogram of Environmental Variables",
       x = "Value",
       y = "Frequency")
```

Create Q-Q plots for each variable to visually assess if the data is normally distributed.

```{r}
par(mfrow=c(2,3))  # set layout for multiple plots

qqnorm(enviro_data_df$n_species, main = "Q-Q Plot: n_species")
qqline(enviro_data_df$n_species, col = "red")

qqnorm(enviro_data_df$n_individuals, main = "Q-Q Plot: n_individuals")
qqline(enviro_data_df$n_individuals, col = "red")

qqnorm(enviro_data_df$elevation, main = "Q-Q Plot: elevation")
qqline(enviro_data_df$elevation, col = "red")

qqnorm(enviro_data_df$np, main = "Q-Q Plot: np")
qqline(enviro_data_df$np, col = "red")

qqnorm(enviro_data_df$ed, main = "Q-Q Plot: ed")
qqline(enviro_data_df$ed, col = "red")

qqnorm(enviro_data_df$pland_mixed_forest, main = "Q-Q Plot: pland_mixed_forest")
qqline(enviro_data_df$pland_mixed_forest, col = "red")

qqnorm(enviro_data_df$ndvi, main = "Q-Q Plot: ndvi")
qqline(enviro_data_df$ndvi, col = "red")
```

Create a model to describe species richness using elevation, edge denity, and ndvi.

```{r}
model <- lm(n_species ~  elevation + ed + ndvi , data=enviro_data_df)

summary(model)
anova(model)
```

Create more diagnostic plots to evaluate the strength of the model.

```{r}
enviro_data_df$yhat.1 <- fitted(model)  
enviro_data_df$resid.1 <- resid(model) 
enviro_data_df$stdresid.1 <- resid(model)/summary(model)$sigma 

plot(yhat.1 ~ n_species, data=enviro_data_df, xlab="Observed n_species", 
     ylab="Predicted n_species", main="Predicted vs. observed n_species",
     pch=19)
abline(a=0,b=1,col="red")   # plot a reference line 


plot(resid.1 ~ yhat.1, data = enviro_data_df, xlab="Predicted n_species", 
     ylab="Residuals",main="Residual Plot",pch=19)
abline(a=0,b=0,col="red")
```

# 7.3 Species Rank Plots {#7.3SpeciesRank}

Species rank plots show the relative abundance (number of individuals) of a species in a community. The number of individuals of each species are sorted in ascending or descending order.

Group the NatureCounts data by species and rank them in order of abundance.

```{r}
species_rank <- mauricie_birds_summary %>%
  filter(!is.na(english_name)) %>%  # Remove rows with NA in english_name
  group_by(english_name) %>%  # Group by species only
  summarize(total_abundance = sum(ObservationCount, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_abundance)) %>%  # Sort in descending order of abundance
  slice_max(total_abundance, n = 40) %>%  # Keep only the top 40 species
  mutate(rank = row_number())  # Assign rank to each species
```

Plot the abundance of each species and its rank across the entire park.

```{r}
ggplot(species_rank, aes(x = reorder(english_name, rank), y = total_abundance)) +
  geom_line(group = 1, size = 1, color = "black") +
  geom_point(size = 2, color = "darkgreen") +
  theme_minimal() +
  labs(
    title = "Abundance of High Rank Species across La Mauricie Park",
    x = "Species",
    y = "Total Abundance"
  ) +
  theme(
    axis.text.x = element_text(size = 8, angle = 60, hjust = 1)
  )
```

# 7.4 Regression Plots {#7.4Regression}

Create scatter plots with regression lines to explore the relationship of species richness with each environmental variable.

```{r}
# Elevation vs. n_species
ggplot(enviro_data_df, aes(x = elevation, y = n_species)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Relationship between Elevation and Species Richness")

# PLAND vs. n_species
ggplot(enviro_data_df, aes(x = pland_mixed_forest, y = n_species)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Relationship between PLAND and Species Richness")

# Edge Density vs. n_species
ggplot(enviro_data_df, aes(x = ed, y = n_species)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Relationship between Edge Density and Species Richness")

# NDVI vs. n_species
ggplot(enviro_data_df, aes(x = ndvi, y = n_species)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Relationship between NDVI and Species Richness")
```

**Conclusion**: There is no clear relationship between species richness and elevation, ED, or NDVI within La Mauricie National Park. **Note**: These results assume that sampling is random with regards to elevation, which is unlikely.

# 7.5 Presence/Absence Plots {#7.5PresenceAbsence}

Create an events matrix containing the geometry for each unique survey event location.

```{r}
events_matrix <- mauricie_birds_df %>%
  dplyr::select(SiteCode, survey_year, survey_month, survey_day, ObservationCount, latitude, longitude) %>%
  distinct() %>%
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326)
```

Assign a point id identifier to each location based on its unique geometry and return the result as a dataframe.

```{r}
events_matrix <- events_matrix %>% 
  group_by(SiteCode, geometry) %>% 
  mutate(point_id = cur_group_id()) %>% 
  ungroup()
```

Restore the longitude and latitude columns, drop the geometry, convert to a dataframe, and keep only the relevant columns.

```{r}
events_matrix <- events_matrix %>%
  mutate(longitude = sf::st_coordinates(.)[,1],  # Extract longitude
         latitude = sf::st_coordinates(.)[,2]) %>%  # Extract latitude
  sf::st_drop_geometry() %>%  # Drop geometry column
  as.data.frame() %>%
  dplyr::select(point_id, SiteCode, survey_year, survey_month, survey_day, longitude, latitude)  # Keep relevant columns
```

Join the environmental variables to the events matrix based on `point_id`.

```{r}
enviro_events_matrix <- left_join(events_matrix, enviro_data_df, by = "point_id")
```

Filter the NatureCounts data for a species of interest and then merge the events matrix with the species-specific data. This will result in NA values for events where the species was not observed.

```{r}
search_species("Golden-crowned Kinglet") #species_id = 14960

kinglet_events <- mauricie_birds_df %>%
  filter(species_id == "14960")

kinglet_events <- left_join(enviro_events_matrix, kinglet_events,
                          by = c("SiteCode",
                               "survey_year",
                               "survey_month",
                               "survey_day",
                               "latitude",
                               "longitude"))
```

Zero-fill for specific species. Here, we replace the NA values in the `ObservationColumn` with 0 and select the column of interest to inspect your data.

```{r}
kinglet_zero_fill <- kinglet_events %>% 
  mutate(ObservationCount = replace(ObservationCount, is.na(ObservationCount), 0)) %>% dplyr::select(point_id, SiteCode, survey_day, survey_month, survey_year, latitude, longitude, elevation:n_individuals, ObservationCount) 
```

Create a Presence/Absence column based on `ObservationCount`.

```{r}
kinglet_presence_absence <- kinglet_zero_fill %>%
  mutate(Presence = ifelse(ObservationCount > 0, 1, 0))  # 1 = Presence, 0 = Absence
```

Summarize the presence/absence data by `point_id`.

```{r}
kinglet_PA_summary <- kinglet_presence_absence %>%
  group_by(point_id) %>%
  summarise(
    total_presence = sum(Presence),  # Total presence
    total_absence = sum(1 - Presence),  # Total absence
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(total_presence, total_absence), 
               names_to = "category", 
               values_to = "count")
```

View the summarized Presence/Absence table.

```{r}
print(kinglet_PA_summary)
```

Pivot the data for plotting.

```{r}
kinglet_PA_summary_long <- PA_summary %>%
  pivot_longer(cols = c(total_presence, total_absence), 
               names_to = "category", 
               values_to = "count")
```

Create a bar plot to summarize the Presence/Absence data, grouped by `survey_year`.  

```{r}
kinglet_PA_summary_year <- kinglet_presence_absence %>%
  group_by(survey_year, category = ifelse(ObservationCount > 0, "Presence", "Absence")) %>%
  summarise(count = n(), .groups = "drop")

ggplot(kinglet_PA_summary_year, aes(x = as.factor(survey_year), y = count, fill = category)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Presence" = "darkgreen", "Absence" = "lightgray")) +
  labs(title = "Golden-crowned Kinglet Presence/Absence per Survey Year",
       x = "Survey Year",
       y = "Number of Observations") +
  theme_minimal()
```
Create a bar plot representing Presence/Absence per `point_id`.

```{r}
ggplot(kinglet_PA_summary_long, aes(x = as.factor(point_id), y = count, fill = category)) +
  geom_col(position = "dodge") +  # Grouped bars
  scale_fill_manual(values = c("total_presence" = "darkgreen", "total_absence" = "lightgray"),
                  labels = c("Absence", "Presence"),
                  name = "Observation Type") +
  labs(title = "Golden-crowned Kinglet Site-Level Presence/Absence Plot",
       x = "Survey Location (point_id)",
       y = "Number of Observations") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),  # Hide x-axis labels
        axis.ticks.x = element_blank(),
        legend.position = "right")
```

\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

Congratulations! You completed Chapter 7 - Summary Tools. In this chapter, you successfully summarized and visualized environmental data, biodiversity counts, and presence/absence data using various plotting techniques . If you'd like to explore some Additional Resources based on previous Spatial Data chapters, the [Raster Tools](Additional%20Resources%20-%20Raster%20Tools.Rmd) article is also available.
