---
title: "AdditionalResources - Raster Summary Tools"
author: "Dimitrios Markou"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Raster Summary Tools 

##### Authors: Dimitrios Markou, Danielle Ethier

This article is a companion to Chapters 4-6 of the Spatial_Data_Tutorial. Here, you will combine and reclassify raster data and perform raster algebra to calculate summary statistics.

**This article expands on Chapters 4-6 of the Spatial Data Tutorial and uses theNational Park boundary and NatureCounts data downloaded in section 4.1 (Data Setup) from [Chapter 4: Elevation Data](04-ElevationData.Rmd). Raster data were downloaded in [Chapter 4: Elevation Data](04-ElevationData.Rmd) and [Chapter 5: Land Cover Data](05-LandcoverData.Rmd) which are also required for this lesson. For quick access, these layers are available for download via the [Google Drive](https://drive.google.com/drive/folders/1gLUC6fROl4kNBvTGselhZif-arPexZbY?usp=sharing) data folders. If you wish to gain experience in how to download, process, and save environmental layers yourself, return to the chapters covered in the Spatial Data series.**

# Learning Objectives

In this article, you will:

-   Transform and combine raster data objects: [Raster Data Transformations](#7.1RasterDataTransformations)

-   Reclassify raster data: [Raster Reclassification](#7.2RasterReclassification)

-   Perform raster algebra using local, global, and zonal functions: [Raster Algebra](#7.3RasterAlgebra)

Load the required packages:

```{r, eval = TRUE, warning = FALSE, message = FALSE}
library(sf)
library(terra)
library(tidyverse)
library(dplyr)
```

Use `st_read()` from the `sf` function to bring in the **mauricie_boundary** which your downloaded in [**Chapter 4: Elevation Data**](04-ElevationData.Rmd) or download via the [Git repository](https://github.com/BirdsCanada/NatureCounts_SpatialData_Tutorial/tree/d83eed714373048c8f9fdb4ea4a17f1fd4bef3f7/Data) or [Google Drive](https://drive.google.com/drive/folders/1gLUC6fROl4kNBvTGselhZif-arPexZbY?usp=sharing) folder and saved to your working directory.

```{r}
mauricie_boundary <- sf::st_read("mauricie_boundary.shp")
```

# Part 1 Raster Data Transformations {#7.1RasterDataTransformations}

Raster data is stored as a SpatRaster object and can have one or multiple layers. To open a SpatRaster, we can use the `rast()` function from the `terra` package. In [Chapter 4: Elevation Data](04-ElevationData.Rmd), you created a mask of La Mauricie National Park using single-layer DTM elevation data. Read this file from your directory and call it `elevation_rast`: 

```{r}
elevation_rast <- rast("path/to/your/dtm_mask.tif")
elevation_rast
```

The function `names()` can be used to rename the layer(s) of a SpatRaster.

```{r}
names(elevation_rast) <- "elevation"
elevation_rast
```

In [Chapter 5: Land Cover Data](05-LandcoverData.Rmd), you created a landcover mask of La Mauricie National Park using the [2015 Land Cover of Canada](https://open.canada.ca/data/en/dataset/4e615eae-b90c-420b-adee-2ca35896caf6) dataset. Read this file from your directory and call it `landcover_rast`:

```{r}
landcover_rast <- rast("path/to/your/landcover_mask.tif")
names(landcover_rast) <- "landcover_class" # Rename the land cover class layer.}
landcover_rast
```

You'll notice that the resolution, extent, and CRS of the elevation and land cover raster layers do not match which prevents us from pulling these layers together. Use the `project()` function from `terra` to transform the coordinate reference system of `elevation_rast` to match that of `landcover_rast` (Canada Atlas Lambert, EPSG: 3979). By default, the "bilinear" method will be used for numeric layers like elevation. Run [help("project")]{.underline} in your **Console** for more details. **This could take about 15 min - stretch break!**

```{r, eval = true, message = FALSE}
elevation_rast_proj <- terra::project(elevation_rast, "EPSG:3979", method = "bilinear")
```

To match the extents of both layers, crop the `landcover_rast` to the extent of the `elevation_rast_proj`.

```{r}
landcover_rast <- crop(landcover_rast, ext(elevation_rast_proj))
```

Resample `landcover_rast` to match the resolution of `elevation_rast_proj` using the **"near"** method. This is the default method used for categorical layers like land cover. Run [help("resample")]{.underline} in your **Console** for more details.

```{r}
landcover_rast_resample <- resample(landcover_rast, elevation_rast_proj, method = "near")
```

Let's verify that our transformations were successful. The `all()` function will help evaluate a set of conditions and produce a TRUE or FALSE result across all our checks for spatial alignment.

```{r}
all(
  crs(landcover_rast_resample) == crs(elevation_rast_proj), # check crs
  ext(landcover_rast_resample) == ext(elevation_rast_proj), # check extent
  res(landcover_rast_resample) == res(elevation_rast_proj) # check resolution
)
```

Let's test plot the layers side by side.

```{r, eval = TRUE, warning = FALSE}
# Match the CRS
mauricie_boundary <- st_transform(mauricie_boundary, crs = st_crs(landcover_rast))

# Set up a 1x2 plotting layout
par(mfrow = c(1, 2))

# Plot elevation
plot(elevation_rast_proj, axes = TRUE, plg = list(title = ""), main = "Elevation (m)")

# Overlay the Mauricie boundary on the first plot
plot(mauricie_boundary, add = TRUE, border = "red", col = NA, lwd = 2)

# Plot landcover
plot(landcover_rast_resample, axes = TRUE, plg = list(title = ""), main = "Land Cover Class")

# Overlay the Mauricie boundary on the second plot
plot(mauricie_boundary, add = TRUE, border = "red", col = NA, lwd = 2)
```

Multiple single-layer SpatRaster with the same CRS, extent, and resolution, can be combined into a single multi-layer SpatRaster with the `c()` function.

```{r}
mauricie_rast <- c(elevation_rast_proj, landcover_rast_resample)
mauricie_rast
```

You can subset a multi-layer SpatRaster with the \$ or [[]] operators.

```{r}
mauricie_rast$elevation

mauricie_rast[["landcover_class"]]
```

# Part 2 Raster Reclassification {#7.2RasterReclassification}

When communicating our data, it might be useful to apply **local** operations such as the reclassification of a range of raster values into a smaller set of classes. In this process, a range of values are binned to a new (categorical) value. Reclassifying rasters is easily achieved using `terra` and the `classify()` function.

First, provide a reclassification table in the form of a matrix which specifies the range of raster values (columns 1 and 2) and the class ID (column 3) like so: ###Could we tell the user what they are reclassifying. That are the meaning of these number?###

```{r}
dtm_rcl_matrix <- matrix(c(0, 200, 1,
                           200, 400, 2,
                           400, 1000, 3), byrow = TRUE, ncol = 3)
```

Then, apply the `classify()` function using the reclassification matrix:

```{r}
elevation_rast_reclass <- classify(elevation_rast_proj, dtm_rcl_matrix)
```

Finally, give meaningful labels to each of the classes: ###Tell the user what this is doing. We are reclassify elevation to low (0-200m), medium (200-400m) and high (400-1000m)?###

```{r, eval = TRUE}
elevation_levels <- data.frame(ID = c(1L, 2L, 3L),
                            category = c("Low",
                                         "Mid",
                                         "High"))
levels(elevation_rast_reclass) <- elevation_levels
```

Visualize the regional DTM and National Park boundary using the reclassification scheme:

```{r, eval = TRUE, warning = FALSE}
# Set up a 1x2 plotting layout
par(mfrow = c(1, 2))  # Set up a 1-row, 2-column plotting layout

# Plot the original DTM mosaic
plot(elevation_rast_proj, main = "Regional DTM")

# Overlay the Mauricie boundary on the first plot
plot(mauricie_boundary, add = TRUE, border = "red", col = NA, lwd = 2)

# Plot the reclassified DTM mosaic
plot(elevation_rast_reclass, main = "Reclassified DTM")

# Overlay the Mauricie boundary on the second plot
plot(mauricie_boundary, add = TRUE, border = "red", col = NA, lwd = 2)

# Reset the plotting layout to the default
par(mfrow = c(1, 1))
```

# Part 3 Raster Algebra (Local, Global, and Zonal Operations) {#7.3RasterAlgebra}

Map algebra comprises a slew of useful operations that can be used to modify or summarize raster cell values with reference to nearby cells, zones, or statistical functions.

**Local** operations involve all cell by cell operations across single or multi-layer rasters. In the [Raster Reclassification](#7.2RasterReclassification) section, we performed a local operation by reclassifying elevation raster values into distinct groups. Custom functions used to calculate spectral indices (see [Chapter 6: SatelliteImagery](06-SatelliteImagery.Rmd)) are another example of local (pixel by pixel) operations.

**Global** operations can be used to derive descriptive raster statistics and summaries across entire SpatRaster. For example, you can visualize your raster data values using a histogram (`hist()`).

```{r, warning = FALSE}
hist(mauricie_rast$elevation, axes = TRUE, 
            main = "Elevation values within La Mauricie National Park",
            xlab = "Elevation (m)",
            col = "Green") 

####Error in h(simpleError(msg, call)) : 
#  error in evaluating the argument 'x' in selecting a method for function 'hist': [subset] invalid name(s)####
```

You can produce global summary statistics for individual SpatRaster layers using the `global()` function. It helps calculate `min()`, `max()`, `mean()`, and `sd()` by specifying the `fun` argument.

```{r}
global_elevation_stats <- global(elevation_rast, fun = c("min", "max", "mean", "sd"), na.rm = TRUE)
print(global_elevation_stats)
```

You can also use custom functions to calculate global raster summary statistics. For example, `my_summary_stats` calculates the min, max, mean, and sd of an input numeric object `x` and returns the summary statistics in a vector (**give this a few minutes**):

```{r}
my_summary_stats <- function(x) {
  out <- c(min = min(x, na.rm = TRUE),
           max = max(x, na.rm = TRUE),
           mean = mean(x, na.rm = TRUE),
           sd = sd(x, na.rm = TRUE))
  return(out)
}
global(elevation_rast, fun = my_summary_stats) # apply the function
```

**Zonal** operations apply an aggregation function to multiple raster cells. Typically, a second (categorical) raster defines the zonal filters (or 'zones'). To calculate the mean elevation per individual land cover class, we can apply the `zonal()` function.

```{r}
zonal_elevation_stats <- zonal(elevation_rast_proj, landcover_rast_resample, fun = "mean", na.rm = TRUE)
zonal_elevation_stats
```